---
title: "In silico simulations (D1b)"
author: "Adrien Taudière"
format: html
---

```{r setup}
#| message: false
library("conflicted")
library("MiscMetabar")
library("targets")
library("here")
library("tibble")
library("tidyr")
library("dplyr")
library("ggplot2")
library("Biostrings")
library("miaSim")

conflicted::conflicts_prefer(dplyr::filter)

here::i_am("In_silico_simulation.qmd")
source(here("config.R"))
devtools::load_all("~/Nextcloud/IdEst/Projets/pqverse/pqverse_pkg/MiscMetabar/")
source("~/Nextcloud/IdEst/Projets/pqverse/pqverse_pkg/comparpq/R/compare_taxo.R")
source("~/Nextcloud/IdEst/Projets/pqverse/pqverse_pkg/comparpq/R/fake_creation.R")
```

# Overview

Two sub-branches following the Bokulich (2020) multi-dataset approach:

- **D1b-A** — InSilicoSeq: real UNITE sequences + simulated Illumina errors → fastq
  → DADA2 → assignment. Ground truth = input fasta taxonomy.
- **D1b-B** — miaSim: neutral (Hubbell) community structure applied to real UNITE
  sequences, bypassing DADA2. Ground truth = input fasta taxonomy. Tests
  assignment accuracy independent of denoising errors.

Both branches reuse `tc_metrics_mock()` machinery from `comparpq`.

---

# D1b-A — InSilicoSeq path

This branch requires InSilicoSeq installed in the `cutadaptenv` conda environment
(or a dedicated env). It is not run inline — follow the steps below, then use the
DADA2 + assignment pipeline on the resulting fastqs.

## Step 1 — Prepare input fasta

Select a curated set of Fungi sequences (50–200 species, spanning ≥ 4 phyla) from
`mini_Unite_Fungi.fasta`. Using the mini DB keeps InSilicoSeq runs fast during
development; switch to the full `Unite_Fungi.fasta` for the publication run.

```{r prepare-iss-input}
# Output path for the curated input fasta
iss_input_fasta <- here("data/data_raw/insilico/iss_input_fungi.fasta")
dir.create(dirname(iss_input_fasta), showWarnings = FALSE, recursive = TRUE)

if (!file.exists(iss_input_fasta)) {
  db_path <- here(
    "data/data_raw/refseq/dada2_format",
    ifelse(mini_db, "mini_Unite_Fungi.fasta", "Unite_Fungi.fasta")
  )
  all_seqs <- Biostrings::readDNAStringSet(db_path)
  set.seed(targets_seed)
  n_iss <- min(100, length(all_seqs))
  picks  <- sample(length(all_seqs), n_iss)
  Biostrings::writeXStringSet(all_seqs[picks], iss_input_fasta)
  message("Wrote ", n_iss, " sequences to ", iss_input_fasta)
}
```

## Step 2 — Run InSilicoSeq (terminal)

Run from the project root in the `cutadaptenv` conda environment. Repeat with
3 different `--seed` values for replicate communities.

```bash
source ~/miniforge3/etc/profile.d/conda.sh && conda activate cutadaptenv

for seed in 1 2 3; do
  iss generate \
    --genomes data/data_raw/insilico/iss_input_fungi.fasta \
    --sequence_type amplicon \
    --n_reads 10000 \
    --abundance zero_inflated_lognormal \
    --model MiSeq \
    --seed ${seed} \
    --output data/data_raw/insilico/iss_seed${seed} \
    --cpus 4
done
```

Outputs: `data/data_raw/insilico/iss_seed{1,2,3}_R{1,2}.fastq.gz`

## Step 3 — Process with existing DADA2 + assignment pipeline

Once fastqs exist, point `script_dada2.R` and `script_assign_taxo_parallel.R` at
`data/data_raw/insilico/` (adapting sample metadata) and run with dedicated stores:

```r
tar_make(script = "script_dada2.R",                store = "store_dada2_insilico")
tar_make(script = "script_assign_taxo_parallel.R", store = "store_assign_taxo_insilico")
```

The truth table for D1b-A is built in the same way as D1b-B below (parse taxonomy
from `iss_input_fasta` headers).

---

# D1b-B — miaSim path

Neutral (Hubbell) community dynamics applied to real UNITE sequences. No DADA2:
each simulated time-step is treated as a perfectly denoised sample.

## Helper — parse UNITE taxonomy from header

```{r parse-taxonomy-fn}
# Extracts the k__/p__/.../s__ taxonomy string from UNITE dada2-format headers.
# Headers: "ID|acc|SH|type|k__Kingdom;p__Phylum;...;s__Species"
parse_unite_taxonomy <- function(headers) {
  tax_str <- sub(".*\\|", "", headers)
  ranks   <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
  prfx    <- c("k__",     "p__",    "c__",   "o__",   "f__",    "g__",   "s__")

  result <- lapply(tax_str, function(h) {
    parts <- strsplit(h, ";")[[1]]
    vapply(seq_along(ranks), function(i) {
      hit <- grep(paste0("^", prfx[i]), parts, value = TRUE)
      if (length(hit) == 0) { return(NA_character_) }
      sub(prfx[i], "", hit[1])
    }, character(1))
  })

  df <- as.data.frame(do.call(rbind, result), stringsAsFactors = FALSE)
  names(df)    <- ranks
  rownames(df) <- headers
  df
}
```

## 1. Sample species from UNITE Fungi

```{r sample-species}
db_path <- here(
  "data/data_raw/refseq/dada2_format",
  ifelse(mini_db, "mini_Unite_Fungi.fasta", "Unite_Fungi.fasta")
)
all_fungi <- Biostrings::readDNAStringSet(db_path)

# Keep only sequences with ≥ genus-level annotation.
has_genus <- !grepl("g__.*Incertae_sedis|g__[^;]+_gen_", names(all_fungi))
all_fungi <- all_fungi[has_genus]

n_species <- min(50, length(all_fungi))
set.seed(targets_seed)
idx       <- sample(length(all_fungi), n_species)
sim_seqs  <- all_fungi[idx]

# Truth table: one row per species, columns = taxonomy ranks.
truth_table <- parse_unite_taxonomy(names(sim_seqs))

cat("Species selected:", n_species, "\n")
cat("Phyla represented:", length(unique(truth_table$Phylum)), "\n")
print(table(truth_table$Phylum))
```

## 2. Simulate community abundances with Hubbell model

Run `n_samples` independent Hubbell simulations (different seeds = independent
communities from the same metacommunity). Each returns a species × time matrix;
we take the final time step as the community composition.

```{r simulate-hubbell}
n_samples <- 10
sim_seed  <- targets_seed

all_counts <- vapply(seq_len(n_samples), function(s) {
  set.seed(sim_seed + s)
  sim <- miaSim::simulateHubbell(
    n_species         = n_species,
    M                 = 200,
    carrying_capacity = 1000,
    k_events          = 50,
    migration_p       = 0.02,
    t_end             = 200
  )
  counts_mat <- SummarizedExperiment::assay(sim, "counts")
  as.integer(counts_mat[, ncol(counts_mat)])
}, integer(n_species))

rownames(all_counts) <- names(sim_seqs)
colnames(all_counts) <- paste0("sim_sample_", seq_len(n_samples))

cat("OTU table dimensions:", nrow(all_counts), "species ×", ncol(all_counts), "samples\n")
cat("Samples with 0 reads:", sum(colSums(all_counts) == 0), "\n")
```

## 3. Build phyloseq + inject fake taxa

```{r build-phyloseq}
pq_sim <- phyloseq::phyloseq(
  phyloseq::otu_table(all_counts, taxa_are_rows = TRUE),
  phyloseq::sample_data(data.frame(
    sample_id = colnames(all_counts),
    row.names = colnames(all_counts)
  ))
)
pq_sim@refseq <- sim_seqs

# Inject shuffled (fake_*) and external (non-Fungi) sequences for TN computation,
# mirroring the real pipeline in script_assign_taxo_parallel.R.
pq_sim <- add_shuffle_seq_pq(pq_sim, prop_fake = prop_fake)
pq_sim <- add_external_seq_pq(
  pq_sim,
  Biostrings::readDNAStringSet(here(fake_ref_fasta))
)

cat("Phyloseq taxa after fake injection:", ntaxa(pq_sim), "\n")
```

## 4. Assign taxonomy

Run a small subset of method × DB combinations. Adjust `sim_values_map` to test
more combinations. The `mini_db` flag from `config.R` controls which database
files are used (mini_* for smoke-testing, full for publication).

```{r assign-taxonomy}
sim_values_map <- tibble::tibble(
  method = c("sintax", "sintax", "dada2"),
  db     = c("Unite_Fungi", "EUK_ITS_v2_Fungi", "Unite_Fungi"),
  min_bootstrap = 0.5
) |>
  dplyr::mutate(
    db_name   = ifelse(mini_db, paste0("mini_", db), db),
    db_path   = paste0(
      ifelse(method == "dada2",
             "data/data_raw/refseq/dada2_format/",
             "data/data_raw/refseq/sintax_format/"),
      db_name, ".fasta"
    ),
    full_name = paste0(method, "__", db, "___", min_bootstrap)
  ) |>
  dplyr::filter(file.exists(here(db_path)))

if (nrow(sim_values_map) == 0) {
  stop("No database files found. Run derive_all_variants() first.")
}

pq_assigned <- pq_sim
for (i in seq_len(nrow(sim_values_map))) {
  row <- sim_values_map[i, ]
  message("Assigning: ", row$full_name)
  pq_assigned <- add_new_taxonomy_pq(
    pq_assigned,
    method        = row$method,
    ref_fasta     = here(row$db_path),
    suffix        = paste0("_", row$full_name),
    min_bootstrap = row$min_bootstrap
  )
}
```

## 5. Compute performance metrics

```{r compute-metrics}
assignment_suffixes <- paste0("_", sim_values_map$full_name)

res_sim <- tc_metrics_mock(
  pq_assigned,
  truth_table,
  assignment_suffixes
)

# Add method / db / bootstrap columns for easier filtering.
res_sim <- res_sim |>
  dplyr::left_join(
    sim_values_map |> dplyr::select(full_name, method, db, min_bootstrap),
    by = c("suffix" = "full_name")
  )

head(res_sim)
```

## 6. Visualise results

### F1 and MCC per taxonomic rank

```{r plot-metrics}
res_sim |>
  dplyr::filter(metric %in% c("F1", "MCC")) |>
  ggplot2::ggplot(ggplot2::aes(
    x    = tax_level,
    y    = value,
    fill = db
  )) +
  ggplot2::geom_col(position = "dodge") +
  ggplot2::facet_grid(metric ~ method, scales = "free_y") +
  ggplot2::labs(
    title = "D1b-B: assignment accuracy on miaSim communities",
    x     = "Taxonomic rank",
    y     = "Score",
    fill  = "Reference DB"
  ) +
  ggplot2::theme_bw()
```

### NA (unclassified) proportion per rank

```{r plot-na}
res_sim |>
  dplyr::filter(metric == "NA_prop") |>
  ggplot2::ggplot(ggplot2::aes(
    x    = tax_level,
    y    = value,
    colour = method,
    group  = full_name
  )) +
  ggplot2::geom_line() +
  ggplot2::geom_point() +
  ggplot2::facet_wrap(~db) +
  ggplot2::labs(
    title = "D1b-B: proportion unclassified per rank",
    x     = "Taxonomic rank",
    y     = "Proportion NA"
  ) +
  ggplot2::theme_bw()
```

---

# Comparison: D1b-B vs mock community (D1)

To compare in silico and mock-community results, load `res_comp_tax` from
`store_assign_taxo` and bind with `res_sim`, then facet by `dataset`.

```{r compare-mock, eval=FALSE}
# Load mock-community metrics from the main store.
res_mock <- targets::tar_read(d_all_taxo, store = here("store_assign_taxo")) |>
  # Reproduce the tc_metrics_mock call from benchmark.qmd ...
  identity()

res_combined <- dplyr::bind_rows(
  dplyr::mutate(res_sim,  dataset = "in_silico_miaSim"),
  dplyr::mutate(res_mock, dataset = "mock_community")
)

res_combined |>
  dplyr::filter(metric == "F1") |>
  ggplot2::ggplot(ggplot2::aes(x = tax_level, y = value, fill = dataset)) +
  ggplot2::geom_boxplot() +
  ggplot2::facet_wrap(~method) +
  ggplot2::theme_bw()
```
