---
title: "Sandbox — distance to the true community"
author: "Adrien Taudière"
---

Work in progress (ROADMAP: "Finalize distance-to-true-community metric").
Moved from the "En chantier" section of the former `benchmark.qmd`; not part
of the project render. Builds, per sample, a presence/absence matrix of the
`Gen_sp_*` calls of every method against the mock truth, then compares them
with Jaccard / Bray-Curtis distances and UpSet plots. Uses
`tidypq::pq_to_tidy()` (decision 13), which reports unassigned ranks as
`"Unknown"` rather than `NA`.

```{r setup}
here::i_am("analysis/sandbox/distance_to_true_community.qmd")  # _quarto.yml would otherwise make here() root = analysis/
source(here::here("analysis/00_setup.R"))
d_all_taxo <- cache_read("d_all_taxo")
taxo_mock  <- cache_read("taxo_mock")
```

## Community matrices per sample

```{r mat-com}
distinct_species_taxo_mock <- distinct(taxo_mock, Gen_sp)

gen_sp_ranks <- grep("^Gen_sp_", names(tax_table_to_df(d_all_taxo)), value = TRUE)

mat_com_interm <- d_all_taxo |>
  pq_to_tidy(ranks = gen_sp_ranks, verbose = FALSE) |>   # filter_zero = TRUE drops abundance == 0
  select(OTU = taxon_id, Sample = sample_id, Abundance = abundance, contains("Gen_sp")) |>
  full_join(tibble(
    "Sample"      = rep("TRUE_sample", nrow(distinct_species_taxo_mock)),
    "Abundance"   = rep(1, nrow(distinct_species_taxo_mock)),
    "Gen_sp_true" = distinct_species_taxo_mock$Gen_sp
  ))

mat_com <- mat_com_interm |>
  pivot_longer(contains("Gen_sp")) |>
  filter(!is.na(value), value != "Unknown") |>
  pivot_wider(names_from = name, values_from = Abundance) |>
  mutate(across(contains("Gen_sp"), ~ tidyr::replace_na(., 0))) |>
  group_by(Sample, value) |>
  summarise(across(contains("Gen_sp"), ~ sum(.)), .groups = "drop")

mat_com_true <- mat_com |>
  filter(Sample == "TRUE_sample") |>
  select(value, Gen_sp_true)

one_sample_matrix <- function(sample_name) {
  mat_com |>
    filter(Sample == sample_name) |>
    group_by(value) |>
    select(contains("Gen_sp")) |>
    mutate(val = across(contains("Gen_sp"), ~ ifelse(. == 0, 0, 1)))
}

mat_comA <- one_sample_matrix("samp_ITSEnz_DikA")
mat_comB <- one_sample_matrix("samp_ITSEnz_DikB")
mat_comC <- one_sample_matrix("samp_ITSEnz_DikC")
```

## Distance between communities

```{r distances}
compar <- full_join(mat_comA[, 1:2], mat_com_true) |>
  mutate(across(everything(), ~ tidyr::replace_na(., 0))) |>
  column_to_rownames("value")

vegan::vegdist(t(compar), binary = TRUE)
vegan::vegdist(t(compar), binary = FALSE)
```

## UpSet plots

```{r upset}
upset_df <- function(mat_com_x) {
  full_join(mat_com_x, mat_com_true, by = join_by(value == value)) |>
    column_to_rownames("value") |>
    select(contains("Gen_sp")) |>
    mutate(across(everything(), ~ ifelse(. == 0, FALSE, TRUE))) |>
    as_tibble() |>
    mutate(across(everything(), ~ tidyr::replace_na(.x, FALSE)))
}

data_upset_df_comA <- upset_df(mat_comA)
data_upset_df_comB <- upset_df(mat_comB)
data_upset_df_comC <- upset_df(mat_comC)

ComplexUpset::upset(data_upset_df_comA, intersect = colnames(data_upset_df_comA), min_size = 2)
ComplexUpset::upset(data_upset_df_comB, intersect = colnames(data_upset_df_comB), min_size = 2)
ComplexUpset::upset(data_upset_df_comC, intersect = colnames(data_upset_df_comC), min_size = 2)
```

## Root mean square error (RMSE)

To do. See also
https://adriancorrendo.github.io/metrica/articles/classification_case.html
(`metrica::confusion_matrix`).
