---
title: "01 — Load, clean, compute metrics"
author: "Adrien Taudière"
---

Reads `d_all_taxo` from `store_assign_taxo`, cleans the taxonomy strings,
builds the `Gen_sp_*` columns and the consensus columns (Q3.1, Q3.3), then
computes the mock-community metrics (`res_comp_tax`, M1.1). Every later
chapter reads its inputs from `analysis/_cache/`, written at the end of this
file.

```{r setup}
here::i_am("analysis/01_load_and_clean.qmd")  # _quarto.yml would otherwise make here() root = analysis/
source(here::here("analysis/00_setup.R"))
```

# Import from store_assign_taxo

```{r load}
d_all_taxo <- tar_read(d_all_taxo, store = store_assign_taxo)
taxo_mock  <- tar_read(taxo_mock,  store = store_assign_taxo) |>
  rename(Gen_sp = Species)
```

## Clean taxonomy strings

```{r clean}
d_all_taxo <- d_all_taxo |>
  taxtab_replace_pattern_by_NA(
    patterns = ".*_sp$",
    progress_bar = TRUE,
    taxonomic_ranks =
      colnames(d_all_taxo@tax_table)[grep("^Species_", colnames(d_all_taxo@tax_table))]
  ) |>
  taxtab_replace_pattern_by_NA(
    patterns = c(
      ".*_incertae_sedis",
      ".*_Incertae_sedis",
      "unclassified.*",
      "^$"
    ),
    progress_bar = TRUE
  ) |>
  simplify_taxo()
```

## Build `Gen_sp_<method_db>` columns

Genus + Species concatenated per assignment column (candidate for
`tidypq::mutate_taxa_pq()`, ROADMAP S5.1).

```{r gen-sp}
taxtab_d_all <- d_all_taxo@tax_table |>
  as.data.frame()

cl_names <- gsub(
  "Kingdom|Phylum|Phyla|Class|Order|Family|Genus|Species", NA,
  unique(gsub("Kingdom_|Phyla_|Phylum_|Class_|Order_|Family_|Genus_|Species_", "",
              colnames(d_all_taxo@tax_table)))
) |>
  na.omit() |>
  as.vector()

cl_names <- grep("Taxa_name_", cl_names, invert = TRUE, value = TRUE)

for (cl in cl_names) {
  if (paste0("Species_", cl) %in% colnames(taxtab_d_all)) {
    sp <- as.vector(taxtab_d_all[, paste0("Species_", cl)])
  } else {
    sp <- NA
  }
  if (paste0("Genus_", cl) %in% colnames(taxtab_d_all)) {
    gen <- as.vector(taxtab_d_all[, paste0("Genus_", cl)])
  } else {
    gen <- NA
  }

  if (sum(grepl("_", sp)) == sum(!is.na(sp))) {
    taxtab_d_all[paste0("Gen_sp_", cl)] <- sp
  } else {
    taxtab_d_all[paste0("Gen_sp_", cl)] <- paste(gen, sp, sep = "_")
  }
  taxtab_d_all[grepl("NA", as.vector(taxtab_d_all[paste0("Gen_sp_", cl)][, 1])), paste0("Gen_sp_", cl)] <- NA
}

d_all_taxo@tax_table <- tax_table(as.matrix(taxtab_d_all))
```

# Consensus columns

Three base strategies, then (Q3.1) `consensus` + `abs_majority`, the
`strict = TRUE` variants of all five, and (Q3.3) `rel_majority` with
`nb_agree_threshold` 2 and 3. `preference_pattern` is set in `00_setup.R`.

```{r consensus-base}
ranks_base <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Gen_sp")

d_all_taxo <- d_all_taxo |>
  resolve_taxo_conflict(
    pattern_tax_ranks = ranks_base,
    method = "rel_majority",
    new_names = paste0(ranks_base, "_rel_majority_consensus"),
    replace_collapsed_rank_by_NA = TRUE
  ) |>
  resolve_taxo_conflict(
    pattern_tax_ranks = ranks_base,
    method = "preference",
    new_names = paste0(ranks_base, "_preference_consensus"),
    preference_pattern = preference_pattern,
    replace_collapsed_rank_by_NA = TRUE
  ) |>
  resolve_taxo_conflict(
    pattern_tax_ranks = ranks_base,
    method = "unanimity",
    new_names = paste0(ranks_base, "_unanimity_consensus"),
    replace_collapsed_rank_by_NA = TRUE
  )
```

```{r consensus-extra}
extra_consensus_configs <- list(
  list(method = "consensus",    strict = FALSE, nb = 1L, suffix = "consensus_consensus"),
  list(method = "abs_majority", strict = FALSE, nb = 1L, suffix = "abs_majority_consensus"),
  list(method = "rel_majority", strict = TRUE,  nb = 1L, suffix = "rel_majority_strict_consensus"),
  list(method = "preference",   strict = TRUE,  nb = 1L, suffix = "preference_strict_consensus"),
  list(method = "unanimity",    strict = TRUE,  nb = 1L, suffix = "unanimity_strict_consensus"),
  list(method = "consensus",    strict = TRUE,  nb = 1L, suffix = "consensus_strict_consensus"),
  list(method = "abs_majority", strict = TRUE,  nb = 1L, suffix = "abs_majority_strict_consensus"),
  list(method = "rel_majority", strict = FALSE, nb = 2L, suffix = "rel_majority_nb2_consensus"),
  list(method = "rel_majority", strict = FALSE, nb = 3L, suffix = "rel_majority_nb3_consensus")
)

for (cfg in extra_consensus_configs) {
  args <- list(
    physeq              = d_all_taxo,
    pattern_tax_ranks   = ranks_base,
    method              = cfg$method,
    strict              = cfg$strict,
    nb_agree_threshold  = cfg$nb,
    new_names           = paste0(ranks_base, "_", cfg$suffix),
    replace_collapsed_rank_by_NA = TRUE
  )
  if (cfg$method == "preference") {
    args$preference_pattern <- preference_pattern
  }
  d_all_taxo <- do.call(resolve_taxo_conflict, args)
}

all_consensus_suffixes <- c(
  "rel_majority_consensus",
  "preference_consensus",
  "unanimity_consensus",
  vapply(extra_consensus_configs, \(cfg) cfg$suffix, character(1))
)
```

# Performance metrics

## Proportion of NA per rank (exploratory)

```{r na-prop}
df_na <- data.frame(
  "method_db" = names(colSums(is.na(d_all_taxo@tax_table)) / ntaxa(d_all_taxo)),
  "prop_na"   = as.vector(colSums(is.na(d_all_taxo@tax_table)) / ntaxa(d_all_taxo))
) |>
  mutate(tax_level = gsub("_.*", "", method_db)) |>
  mutate(tax_level = substr(tax_level, 1, 1)) |>
  filter(tax_level %in% c("K", "P", "C", "O", "F", "G", "S")) |>
  mutate(bootstrap = gsub(".*__", "", method_db)) |>
  mutate(db = sub(".*__", "", gsub("___.*", "", method_db))) |>
  mutate(algo = dplyr::case_when(
    grepl("dada2", method_db)                  ~ "dada2",
    grepl("sintax", method_db)                 ~ "sintax",
    grepl("lca", method_db)                    ~ "lca",
    grepl("blastn", method_db)                 ~ "blastn",
    grepl("rel_majority_consensus", method_db) ~ "rel_majority_consensus",
    grepl("preference_consensus", method_db)   ~ "preference_consensus",
    grepl("unanimity_consensus", method_db)    ~ "unanimity_consensus",
    TRUE                                       ~ method_db
  )) |>
  mutate(
    db        = ifelse(grepl("consensus", algo), NA, db),
    bootstrap = ifelse(grepl("consensus", algo), NA, bootstrap)
  )

prop_fake_taxa <- sum(grepl("external|fake", taxa_names(d_all_taxo))) / ntaxa(d_all_taxo)

df_na |>
  filter(algo %in% c("unanimity_consensus", "preference_consensus", "rel_majority_consensus")) |>
  ggplot() +
  geom_bar(aes(x = prop_na, y = db, fill = db), stat = "identity", position = position_dodge()) +
  geom_vline(xintercept = prop_fake_taxa) +
  facet_grid(algo ~ factor(tax_level, levels = c("K", "P", "C", "O", "F", "G", "S"))) +
  xlim(c(0, 1)) + no_legend()

df_na |>
  filter(algo %in% single_methods) |>
  ggplot() +
  geom_bar(aes(x = prop_na, y = db, fill = db), stat = "identity", position = position_dodge()) +
  geom_vline(xintercept = prop_fake_taxa) +
  facet_grid(algo ~ factor(tax_level, levels = c("K", "P", "C", "O", "F", "G", "S"))) +
  xlim(c(0, 1)) + no_legend()
```

## Rank columns of interest

```{r ranks-df}
ranks_df_list <- list()
for (db_name in c(all_consensus_suffixes, values_map$full_name)) {
  ranks_df_list[[db_name]] <- paste0(ranks_base, "_", db_name)
}
ranks_df <- do.call(data.frame, ranks_df_list)

missing_cols <- setdiff(as.vector(unlist(ranks_df)), colnames(d_all_taxo@tax_table))
if (length(missing_cols) > 0) {
  print(missing_cols)
  stop("Columns expected from values_map / consensus suffixes are missing in d_all_taxo.")
}
```

## `tc_metrics_mock()` and M1.1 enrichment

```{r metrics}
res_comp_tax <- tc_metrics_mock(
  d_all_taxo,
  ranks_df = ranks_df,
  true_values_df = taxo_mock
)

res_comp_tax <- res_comp_tax |>
  mutate(bootstrap = gsub(".*__", "", method_db)) |>
  mutate(algo = gsub("__.*", "", method_db)) |>
  mutate(db = gsub(".*__", "", gsub("___.*", "", method_db)))

# M1.1: vote_algorithm, bootstrap_num, TAR/TDR aliases
# (TAR = PPV = TP/(TP+FP); TDR = TPR = TP/(TP+FN)).
res_comp_tax <- res_comp_tax |>
  mutate(
    vote_algorithm = ifelse(
      algo == "blastn",
      stringr::str_extract(method_db, "(?<=\\.\\.\\.)[a-z_]+(?=\\.\\.\\.)"),
      NA_character_
    ),
    bootstrap_num = suppressWarnings(
      as.numeric(stringr::str_extract(bootstrap, "^[0-9.]+"))
    )
  )

res_comp_tax <- dplyr::bind_rows(
  res_comp_tax,
  res_comp_tax |> filter(metrics == "PPV") |> mutate(metrics = "TAR"),
  res_comp_tax |> filter(metrics == "TPR") |> mutate(metrics = "TDR")
)

# Consensus-row metadata: is_consensus, strict flag, nb_agree_threshold value.
res_comp_tax <- res_comp_tax |>
  mutate(
    is_consensus     = !algo %in% single_methods,
    consensus_strict = is_consensus & grepl("_strict_", method_db),
    nb_agree         = dplyr::case_when(
      grepl("_nb2_", method_db) ~ 2L,
      grepl("_nb3_", method_db) ~ 3L,
      TRUE ~ 1L
    )
  )
```

# Hand over to the next chapters

```{r cache}
cache_write(d_all_taxo,   "d_all_taxo")
cache_write(taxo_mock,    "taxo_mock")
cache_write(res_comp_tax, "res_comp_tax")
cache_write(ranks_df,     "ranks_df")
```

# Exploratory views (not in the manuscript)

```{r explore-metrics}
plot_tc_metrics_mock(res_comp_tax, metric = "ACC", modality_x = "tax_level") + no_legend()

plot_tc_metrics_mock(res_comp_tax, modality_x = "algo", modality_line = "metrics",
                     modality_column = "tax_level", modality_fill = "db")

for (m in c("FP", "MCC", "F1_score", "ACC")) {
  print(
    plot_tc_metrics_mock(res_comp_tax, metric = m, modality_x = "db", modality_line = "algo",
                         modality_column = "tax_level", modality_fill = "bootstrap") +
      coord_flip() + geom_hline(yintercept = 1) + geom_hline(yintercept = 0, linewidth = 0.2)
  )
}
```

```{r explore-per-db}
res_comp_tax |>
  filter(metrics %in% c("ACC", "F1_score", "MCC")) |>
  filter(bootstrap == 0.5 | grepl("consensus", algo) | algo == "blastn") |>
  filter(tax_level %in% c("Class", "Order", "Family", "Genus", "Gen_sp")) |>
  plot_tc_metrics_mock(modality_x = "tax_level", modality_line = "metrics",
                       modality_column = "algo", modality_fill = "db")

for (one_db in c("Unite", "Unite_Fungi", "EUK_ITS_v2")) {
  print(
    res_comp_tax |>
      filter(metrics %in% c("ACC", "F1_score", "MCC", "FP", "FN")) |>
      filter(tax_level %in% c("Class", "Order", "Family", "Genus", "Gen_sp")) |>
      filter(db == one_db) |>
      plot_tc_metrics_mock(modality_x = "bootstrap", modality_line = "metrics",
                           modality_column = "tax_level", modality_fill = "algo") +
      ggtitle(one_db)
  )
}
```

```{r explore-tpr-tnr}
res_comp_tax |>
  filter(metrics %in% c("TPR", "TNR")) |>
  mutate(db_unite = ifelse(grepl("Unite", method_db), "Unite", "Euk")) |>
  filter(bootstrap == 0.5) |>
  filter(!tax_level %in% c("Kingdom", "Phylum", "Class", "Gen_sp")) |>
  pivot_wider(names_from = metrics, values_from = values) |>
  ggplot(aes(x = TPR, y = TNR, pch = algo, color = db)) +
  geom_point(size = 3, alpha = 0.8) +
  xlim(c(0.4, 1)) + ylim(c(0.4, 1)) +
  facet_wrap(db_unite ~ tax_level)

p <- res_comp_tax |>
  filter(metrics %in% c("TPR", "TNR")) |>
  group_by(method_db) |>
  pivot_wider(names_from = metrics, values_from = values) |>
  summarise(
    TPR_mean = mean(TPR),
    TNR_mean = mean(TNR),
    TPR_sd = var(TPR),
    TNR_sd = var(TNR)
  ) |>
  mutate(bootstrap = gsub(".*__", "", method_db)) |>
  mutate(algo = gsub("__.*", "", method_db)) |>
  mutate(db = gsub(".*__", "", gsub("___.*", "", method_db))) |>
  mutate(db_unite = ifelse(grepl("Unite", method_db), "Unite", "Euk")) |>
  ggplot(aes(x = TPR_mean, y = TNR_mean, color = algo)) +
  geom_point()

p +
  geom_segment(aes(x = TPR_mean, y = TNR_mean - TNR_sd, yend = TNR_mean + TNR_sd), alpha = 0.5) +
  geom_segment(aes(y = TNR_mean, x = TPR_mean - TPR_sd, xend = TPR_mean + TPR_sd), alpha = 0.5) +
  ggrepel::geom_label_repel(
    data = subset(p$data, grepl("consensus", method_db) | TNR_mean > 0.85 | TPR_mean > 0.85),
    aes(label = method_db), size = 2, max.overlaps = 10, force = 3
  ) +
  xlim(c(0.5, 1)) + ylim(c(0.5, 1)) + facet_wrap(db_unite ~ .)
```

```{r explore-tc-bar}
tc_bar(d_all_taxo,
       rank_1 = paste0("Family_", preference_pattern),
       rank_2 = "Family_sintax__Unite___0.5",
       color_rank = paste0("Genus_", preference_pattern))

tc_circle(d_all_taxo,
          "Family_sintax__Unite___0.5",
          paste0("Family_", preference_pattern))
```
