---
title: "01 — Load, clean, score against the per-unit truth"
author: "Adrien Taudière"
---

Reads the assignments of every dataset that has a store, cleans the taxonomy
strings, builds the `Gen_sp_*` and the consensus columns (Q3.1, Q3.3), gives
every unit its own truth (`R/mock_truth.R`, ROADMAP 0.6) and scores the
columns with `comparpq::tc_metrics_unit()` (ROADMAP 0.7). Every later chapter
reads `res_comp_tax` from `analysis/_cache/`, written at the end of this file.

Three inputs are scored side by side, told apart by the `unit` column:

| `unit` | Object | What it is |
|---|---|---|
| `ASV` | `d_all_taxo` | the ASVs, raw and ITSx-trimmed (`itsx_*` columns), plus the consensus columns |
| `OTU` | `d_all_taxo_otu` | the 97 % OTUs (`d_vs`) assigned directly (`otu_*` columns, OD 4) |
| `OTU_derived` | `d_otu_taxo` | the same clusters, but their taxonomy resolved from the member ASVs (decision 25) |

`OTU` against `OTU_derived` is the "cluster then assign" versus "assign then
cluster" comparison of Q5; both are kept on the developer's decision of
2026-09-22.

```{r setup}
here::i_am("analysis/01_load_and_clean.qmd")  # _quarto.yml would otherwise make here() root = analysis/
source(here::here("analysis/00_setup.R"))
source(here("R/mock_truth.R"))
source(here("R/otu_taxonomy.R"))
source(here("R/score_assignments.R"))
```

# Datasets

One row per dataset whose assignment store exists (`analysis_datasets` in
`00_setup.R`). The truth is built from the DADA2 store of the dataset, never
from the assignment inputs: the ITSx input carries the same taxa as the raw
ASVs but trimmed sequences, which must not be re-matched against the Sanger
reads.

```{r datasets}
analysis_datasets
```

# Helpers

```{r helpers-clean}
# Same cleaning for every object: "<Genus>_sp" species, the incertae sedis and
# unclassified labels and the empty strings become NA.
clean_taxo <- function(physeq) {
  species_cols <- grep(
    "^Species_",
    colnames(physeq@tax_table),
    value = TRUE
  )
  physeq |>
    taxtab_replace_pattern_by_NA(
      patterns = ".*_sp$",
      progress_bar = FALSE,
      taxonomic_ranks = species_cols
    ) |>
    taxtab_replace_pattern_by_NA(
      patterns = c(
        ".*_incertae_sedis",
        ".*_Incertae_sedis",
        "unclassified.*",
        "^$"
      ),
      progress_bar = FALSE
    ) |>
    simplify_taxo()
}
```

```{r helpers-gen-sp}
# Genus + Species concatenated per assignment column. The tax_table is read
# with `tidypq::tax_table_to_df()` and the new columns are written back in one
# `tidypq::mutate_taxa_pq()` call (decision 13).
assignment_labels <- function(taxtab_df) {
  labels <- gsub(
    "Kingdom|Phylum|Phyla|Class|Order|Family|Genus|Species", NA,
    unique(gsub("Kingdom_|Phyla_|Phylum_|Class_|Order_|Family_|Genus_|Species_", "",
                setdiff(names(taxtab_df), "taxon")))
  ) |>
    na.omit() |>
    as.vector()
  grep("Taxa_name_", labels, invert = TRUE, value = TRUE)
}

# Species when it already carries the genus (Genus_species), else Genus_Species;
# any "NA" left by the paste becomes a real NA.
gen_sp_for <- function(tt, cl) {
  sp  <- if (paste0("Species_", cl) %in% names(tt)) tt[[paste0("Species_", cl)]] else rep(NA_character_, nrow(tt))
  gen <- if (paste0("Genus_", cl) %in% names(tt)) tt[[paste0("Genus_", cl)]] else rep(NA_character_, nrow(tt))
  out <- if (sum(grepl("_", sp)) == sum(!is.na(sp))) sp else paste(gen, sp, sep = "_")
  out[grepl("NA", out)] <- NA_character_
  out
}

add_gen_sp <- function(physeq) {
  taxtab_df <- tax_table_to_df(physeq, convert = FALSE)
  cl_names <- assignment_labels(taxtab_df)
  gen_sp_cols <- setNames(
    lapply(cl_names, \(cl) gen_sp_for(taxtab_df, cl)),
    paste0("Gen_sp_", cl_names)
  )
  do.call(mutate_taxa_pq, c(list(physeq = physeq), gen_sp_cols))
}
```

```{r helpers-ranks-df}
# The `ranks_df` that tc_metrics_unit() takes: one row per rank of
# `truth_ranks` (its Species row is the `Gen_sp_*` column, because the truth
# holds the underscored binomial and EUKARYOME stores the epithet alone), one
# column per assignment. Assignments with no column at all are dropped on the
# mini stores, where blastn can find no hit in a mini_* database and
# combine_taxo_assignments(allow_empty = TRUE) adds no column; in production an
# empty assignment is a failure.
build_ranks_df <- function(physeq, assignments, ranks = tax_order) {
  ranks_df <- as.data.frame(setNames(
    lapply(assignments, \(a) paste0(ranks, "_", a)),
    assignments
  ))
  present <- names(tax_table_to_df(physeq, convert = FALSE))
  missing_cols <- setdiff(as.vector(unlist(ranks_df)), present)
  empty <- names(ranks_df)[vapply(ranks_df, \(cols) all(cols %in% missing_cols), logical(1))]
  if (analysis_mini && all(missing_cols %in% unlist(ranks_df[empty]))) {
    if (length(empty) > 0) {
      message("mini stores: ", length(empty), " empty assignments dropped: ",
              paste(head(empty, 5), collapse = ", "),
              if (length(empty) > 5) ", ...")
    }
    ranks_df <- ranks_df[setdiff(names(ranks_df), empty)]
  } else if (length(missing_cols) > 0) {
    print(head(missing_cols, 20))
    stop("Columns expected from values_map / consensus suffixes are missing.")
  }
  ranks_df
}
```

# Consensus columns

Three base strategies, then (Q3.1) `consensus` + `abs_majority`, the
`strict = TRUE` variants of all five, and (Q3.3) `rel_majority` with
`nb_agree_threshold` 2 and 3. `preference_pattern` is set in `00_setup.R`.

Q3 votes over one column per method × database, at the default settings
(developer, 2026-09-22): 4 × 9 = 36 columns of equal weight. Since the grid of
2026-09-17 a method contributes as many columns as it has parameter rows, so
the 12 blastn rows per database would carry 57 % of the vote and count the same
BLAST search four times. Every other column — the other parameter values, and
the ITSx rows of Q4, which `preference_pattern` would match too — is set aside
before the vote and added back after.

```{r helpers-consensus}
extra_consensus_configs <- list(
  list(method = "consensus",    strict = FALSE, nb = 1L, suffix = "consensus_consensus"),
  list(method = "abs_majority", strict = FALSE, nb = 1L, suffix = "abs_majority_consensus"),
  list(method = "rel_majority", strict = TRUE,  nb = 1L, suffix = "rel_majority_strict_consensus"),
  list(method = "preference",   strict = TRUE,  nb = 1L, suffix = "preference_strict_consensus"),
  list(method = "unanimity",    strict = TRUE,  nb = 1L, suffix = "unanimity_strict_consensus"),
  list(method = "consensus",    strict = TRUE,  nb = 1L, suffix = "consensus_strict_consensus"),
  list(method = "abs_majority", strict = TRUE,  nb = 1L, suffix = "abs_majority_strict_consensus"),
  list(method = "rel_majority", strict = FALSE, nb = 2L, suffix = "rel_majority_nb2_consensus"),
  list(method = "rel_majority", strict = FALSE, nb = 3L, suffix = "rel_majority_nb3_consensus")
)

all_consensus_suffixes <- c(
  "rel_majority_consensus",
  "preference_consensus",
  "unanimity_consensus",
  vapply(extra_consensus_configs, \(cfg) cfg$suffix, character(1))
)

# The default settings of filter_default_settings() (00_setup.R), as the
# `full_name` of one input: one column per method and database.
default_names_for <- function(input = "none") {
  values_map |>
    filter(preprocess == input) |>
    filter(
      (method %in% c("dada2", "sintax") & min_bootstrap == 0.5) |
        (method == "lca" & lca_cutoff == 1) |
        (method == "blastn" & vote_algorithm == "rel_majority" & min_id == 95)
    ) |>
    pull(full_name)
}

# Columns the Q3 consensus votes over.
consensus_names <- default_names_for("none")

add_consensus_columns <- function(physeq, ranks = tax_order) {
  consensus_cols <- as.vector(outer(ranks, consensus_names, paste, sep = "_"))
  assignment_cols <- grep(
    paste0("^(", paste(ranks, collapse = "|"), ")_"),
    colnames(physeq@tax_table),
    value = TRUE
  )
  missing_consensus <- setdiff(consensus_cols, assignment_cols)
  if (length(missing_consensus) > 0 && !analysis_mini) {
    print(missing_consensus)
    stop("Default-settings columns expected by the Q3 consensus are missing.")
  }
  consensus_cols <- intersect(consensus_cols, assignment_cols)
  aside_cols <- setdiff(assignment_cols, consensus_cols)
  aside_taxtab <- tax_table_to_df(physeq, convert = FALSE)[, c("taxon", aside_cols)]
  message(
    "Q3 consensus over ", length(consensus_cols) / length(ranks),
    " assignment columns per rank; ", length(aside_cols), " columns set aside"
  )
  if (length(aside_cols) > 0) {
    physeq <- select_taxa_pq(physeq, !dplyr::all_of(aside_cols))
  }

  base_configs <- list(
    list(method = "rel_majority", strict = FALSE, nb = 1L, suffix = "rel_majority_consensus"),
    list(method = "preference",   strict = FALSE, nb = 1L, suffix = "preference_consensus"),
    list(method = "unanimity",    strict = FALSE, nb = 1L, suffix = "unanimity_consensus")
  )
  for (cfg in c(base_configs, extra_consensus_configs)) {
    args <- list(
      physeq              = physeq,
      pattern_tax_ranks   = ranks,
      method              = cfg$method,
      strict              = cfg$strict,
      nb_agree_threshold  = cfg$nb,
      new_names           = paste0(ranks, "_", cfg$suffix),
      replace_collapsed_rank_by_NA = TRUE
    )
    if (cfg$method == "preference") {
      args$preference_pattern <- preference_pattern
    }
    physeq <- do.call(resolve_taxo_conflict, args)
  }

  if (length(aside_cols) > 0) {
    aside_rows <- match(taxa_names(physeq), aside_taxtab$taxon)
    physeq <- do.call(
      mutate_taxa_pq,
      c(list(physeq = physeq), as.list(aside_taxtab[aside_rows, aside_cols, drop = FALSE]))
    )
  }
  physeq
}
```

# One dataset at a time

```{r prepare}
# Everything one dataset needs to be scored: the three objects, their truths
# and the lineage of the external controls.
prepare_dataset <- function(ds) {
  message("\n== ", ds$dataset, " ==")
  d_all_taxo <- tar_read(d_all_taxo, store = ds$store) |>
    clean_taxo() |>
    add_gen_sp()
  d_all_taxo_otu <- tar_read(d_all_taxo_otu, store = ds$store) |>
    clean_taxo() |>
    add_gen_sp()

  # Taxonomy of the clusters resolved from the member ASVs (Q5, decision 25):
  # no assignment is recomputed, an OTU keeps the value all its assigned member
  # ASVs agree on. Built before the consensus columns, which are an ASV object.
  otu_clusters <- tar_read(otu_clusters, store = ds$store)
  d_otu_taxo <- resolve_otu_taxonomy(d_all_taxo, otu_clusters, method = "unanimity")
  message(
    ntaxa(d_all_taxo), " taxa (ASVs and controls) merged into ",
    ntaxa(d_otu_taxo), " (OTUs and controls)"
  )

  d_all_taxo <- add_consensus_columns(d_all_taxo)

  cfg <- mock_sanger_tables[mock_sanger_tables$dataset == ds$dataset, ]
  truth_asv <- build_mock_truth(tar_read(d_asv, store = ds$dada2_store), cfg)
  truth_otu <- build_mock_truth(tar_read(d_vs, store = ds$dada2_store), cfg)
  message(
    sum(!is.na(truth_asv$truth_depth)), " of ", nrow(truth_asv),
    " ASVs and ", sum(!is.na(truth_otu$truth_depth)), " of ", nrow(truth_otu),
    " OTUs have a truth"
  )

  list(
    dataset = ds$dataset,
    d_all_taxo = d_all_taxo,
    d_all_taxo_otu = d_all_taxo_otu,
    d_otu_taxo = d_otu_taxo,
    truth_asv = truth_asv,
    truth_otu = truth_otu,
    external_truth = external_control_truth(taxa_names(d_all_taxo))
  )
}

prepared <- setNames(
  lapply(seq_len(nrow(analysis_datasets)), \(i) prepare_dataset(analysis_datasets[i, ])),
  analysis_datasets$dataset
)
```

# Scoring

`score_assignments()` (`R/score_assignments.R`) groups the columns by the
treatment their database imposes on the `external_` controls — in the matrix on
the three `_Fungi` databases, aside on the whole releases and on the three
`_Fungi_rep` (with `ext_correct` at the kingdom only, OD 6, 15, 22) — and calls
`comparpq::tc_metrics_unit()` once per group. The consensus columns have no
database and are scored both ways; the `external_scoring` column tells the two
readings apart.

```{r score}
asv_names <- values_map$full_name[values_map$preprocess %in% c("none", "itsx")]
otu_names <- values_map$full_name[values_map$preprocess == "otu"]

score_dataset <- function(prep) {
  asv <- build_ranks_df(prep$d_all_taxo, c(all_consensus_suffixes, asv_names))
  otu <- build_ranks_df(prep$d_all_taxo_otu, otu_names)
  derived <- build_ranks_df(prep$d_otu_taxo, asv_names)

  bind_rows(
    score_assignments(
      prep$d_all_taxo, asv, prep$truth_asv, db_meta,
      external_truth = prep$external_truth, truth_ranks = truth_ranks
    ) |>
      mutate(unit = "ASV"),
    score_assignments(
      prep$d_all_taxo_otu, otu, prep$truth_otu, db_meta,
      external_truth = prep$external_truth, truth_ranks = truth_ranks
    ) |>
      mutate(unit = "OTU"),
    score_assignments(
      prep$d_otu_taxo, derived, prep$truth_asv, db_meta,
      external_truth = prep$external_truth, truth_ranks = truth_ranks
    ) |>
      mutate(unit = "OTU_derived")
  ) |>
    mutate(dataset = prep$dataset)
}

res_comp_tax <- bind_rows(lapply(prepared, score_dataset)) |>
  # tc_metrics_unit() labels the last rank after the truth table ("Species");
  # the columns it read are the `Gen_sp_*` ones, and every chapter filters on
  # "Gen_sp" (tax_order).
  mutate(tax_level = ifelse(tax_level == "Species", "Gen_sp", tax_level)) |>
  enrich_metrics()

res_comp_tax |>
  count(dataset, unit, external_scoring) |>
  arrange(dataset, unit)
```

# M1.5 — abundance filter

Hleap et al. 2021 dropped the ASVs below 8 reads before assigning them. Here
the filter is applied afterwards, as a sweep: the metrics are recomputed on the
units carrying at least *n* reads, *n* running from 1 (no filter) to 100. It
answers whether the ranking of the methods and databases survives the filter,
and how much of each score rests on units nobody would keep.

The sweep runs on the default settings only (the four methods × the nine
databases, `consensus_names`) and on the ASV and OTU inputs: it multiplies the
scoring by the number of thresholds, and the parameter sweeps of Q1 are read at
one threshold.

**The filter applies to the real units only.** The negative controls are
injected sequences and carry no reads at all (`add_shuffle_seq_pq()` and
`add_external_seq_pq()` add them to the tax_table and the refseq, with a zero
count), so filtering them with the rest would remove every one of them at the
first threshold and empty the TN column — TNR would jump from its real value to
0/0 between *n* = 1 and *n* = 2 and read as an effect of the filter.

```{r abundance-sweep}
# 1 is the unfiltered baseline, added to M1.5's list. On the Pauvert mock it
# gives the same counts as 2, because no ASV carries a single read (the
# DADA2 minimum there is 2); on another dataset the two will differ.
abundance_thresholds <- c(1:10, 12, 15, 20, 30, 40, 50, 100)

sweep_abundance <- function(prep, thresholds = abundance_thresholds) {
  inputs <- list(
    ASV = list(
      physeq = prep$d_all_taxo,
      truth = prep$truth_asv,
      names = default_names_for("none")
    ),
    OTU = list(
      physeq = prep$d_all_taxo_otu,
      truth = prep$truth_otu,
      names = default_names_for("otu")
    )
  )
  bind_rows(lapply(names(inputs), function(unit) {
    input <- inputs[[unit]]
    ranks_df <- build_ranks_df(input$physeq, input$names)
    is_control <- grepl(control_pattern, taxa_names(input$physeq))
    bind_rows(lapply(thresholds, function(n) {
      kept <- prune_taxa(taxa_sums(input$physeq) >= n | is_control, input$physeq)
      score_assignments(
        kept, ranks_df, input$truth, db_meta,
        external_truth = prep$external_truth, truth_ranks = truth_ranks
      ) |>
        mutate(
          unit = unit,
          min_reads = n,
          n_units_kept = sum(taxa_sums(input$physeq) >= n & !is_control)
        )
    }))
  })) |>
    mutate(dataset = prep$dataset)
}

res_abundance <- bind_rows(lapply(prepared, sweep_abundance)) |>
  mutate(tax_level = ifelse(tax_level == "Species", "Gen_sp", tax_level)) |>
  enrich_metrics()

res_abundance |>
  distinct(dataset, unit, min_reads, n_units_kept) |>
  tidyr::pivot_wider(names_from = unit, values_from = n_units_kept)
```

```{r abundance-figure}
res_abundance |>
  filter(dataset == analysis_datasets$dataset[[1]]) |>
  filter(metrics %in% c("F1_score", "TPR", "NA_real")) |>
  filter(tax_level %in% c("Genus", "Gen_sp")) |>
  mutate(algo = sub("^otu_", "", algo)) |>
  left_join(select(db_meta, db, simplification), by = "db") |>
  ggplot(aes(x = min_reads, y = values, colour = algo, linetype = unit)) +
  geom_line(aes(group = interaction(method_db, unit)), alpha = 0.8) +
  scale_x_log10() +
  facet_grid(metrics ~ tax_level + simplification) +
  labs(x = "Minimum reads per unit (log scale)", y = NULL)
```

# Hand over to the next chapters

```{r cache}
cache_write(res_comp_tax, "res_comp_tax")
cache_write(res_abundance, "res_abundance")
# `prepared` (the cleaned objects and their truths) is NOT cached: nothing
# outside this file reads it, and on the production stores its three
# phyloseq objects per dataset carry about 1 200 columns each.
```

# Checks

The species row is the one that fails silently: `truth$Species` is the
underscored binomial of `binomial_name()` and the scored column is
`Gen_sp_*`, built from `Genus_*` and `Species_*`. If the two spellings
disagreed, every unit would be an FP and read as a result.

```{r check-species}
first <- prepared[[1]]
best_gen_sp <- res_comp_tax |>
  filter(dataset == first$dataset, unit == "ASV", tax_level == "Gen_sp",
         metrics == "TP", algo %in% single_methods) |>
  slice_max(values, n = 3)
best_gen_sp |> select(method_db, values)
stopifnot(max(best_gen_sp$values) > 0)

# A handful of matching spellings, to read with the eye.
tt_first <- tax_table_to_df(first$d_all_taxo, convert = FALSE)
gen_sp_col <- paste0("Gen_sp_", best_gen_sp$method_db[[1]])
tibble(
  taxon = tt_first$taxon,
  truth = first$truth_asv$Species[match(tt_first$taxon, first$truth_asv$taxon)],
  assigned = tt_first[[gen_sp_col]]
) |>
  filter(!is.na(truth), !is.na(assigned)) |>
  head(8)
```

Every scored unit falls in exactly one cell of the confusion matrix, so the
four cells sum to the number of scored units — constant across the columns of
one scoring group.

```{r check-cells}
res_comp_tax |>
  filter(metrics %in% c("TP", "FP", "FN", "TN")) |>
  group_by(dataset, unit, external_scoring, tax_level, method_db) |>
  summarise(n_units = sum(values), .groups = "drop") |>
  count(dataset, unit, external_scoring, tax_level, n_units) |>
  filter(tax_level %in% c("Genus", "Gen_sp"))
```

# Exploratory views (not in the manuscript)

Everything below reads the first dataset only.

```{r na-prop}
df_na <- tax_table_to_df(first$d_all_taxo, convert = FALSE) |>
  select(-taxon) |>
  summarise(across(everything(), ~ mean(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "method_db", values_to = "prop_na") |>
  mutate(tax_level = gsub("_.*", "", method_db)) |>
  mutate(tax_level = substr(tax_level, 1, 1)) |>
  filter(tax_level %in% c("K", "P", "C", "O", "F", "G", "S")) |>
  mutate(bootstrap = gsub(".*__", "", method_db)) |>
  mutate(db = sub(".*__", "", gsub("___.*", "", method_db))) |>
  mutate(algo = dplyr::case_when(
    grepl("itsx_dada2", method_db)             ~ "itsx_dada2",
    grepl("itsx_sintax", method_db)            ~ "itsx_sintax",
    grepl("itsx_lca", method_db)               ~ "itsx_lca",
    grepl("itsx_blastn", method_db)            ~ "itsx_blastn",
    grepl("dada2", method_db)                  ~ "dada2",
    grepl("sintax", method_db)                 ~ "sintax",
    grepl("lca", method_db)                    ~ "lca",
    grepl("blastn", method_db)                 ~ "blastn",
    grepl("rel_majority_consensus", method_db) ~ "rel_majority_consensus",
    grepl("preference_consensus", method_db)   ~ "preference_consensus",
    grepl("unanimity_consensus", method_db)    ~ "unanimity_consensus",
    TRUE                                       ~ method_db
  )) |>
  mutate(
    db        = ifelse(grepl("consensus", algo), NA, db),
    bootstrap = ifelse(grepl("consensus", algo), NA, bootstrap)
  )

prop_fake_taxa <- sum(grepl("external|fake", taxa_names(first$d_all_taxo))) /
  ntaxa(first$d_all_taxo)

df_na |>
  filter(algo %in% c("unanimity_consensus", "preference_consensus", "rel_majority_consensus")) |>
  ggplot() +
  geom_bar(aes(x = prop_na, y = db, fill = db), stat = "identity", position = position_dodge()) +
  geom_vline(xintercept = prop_fake_taxa) +
  facet_grid(algo ~ factor(tax_level, levels = c("K", "P", "C", "O", "F", "G", "S"))) +
  xlim(c(0, 1)) + no_legend()

df_na |>
  filter(algo %in% single_methods) |>
  ggplot() +
  geom_bar(aes(x = prop_na, y = db, fill = db), stat = "identity", position = position_dodge()) +
  geom_vline(xintercept = prop_fake_taxa) +
  facet_grid(algo ~ factor(tax_level, levels = c("K", "P", "C", "O", "F", "G", "S"))) +
  xlim(c(0, 1)) + no_legend()
```

```{r explore-metrics}
res_first <- filter(res_comp_tax, dataset == first$dataset, unit == "ASV")

plot_tc_metrics_mock(res_first, metric = "ACC", modality_x = "tax_level") + no_legend()

for (m in c("FP", "MCC", "F1_score", "NA_real")) {
  print(
    res_first |>
      filter_default_settings() |>
      plot_tc_metrics_mock(metric = m, modality_x = "db", modality_line = "algo",
                           modality_column = "tax_level", modality_fill = "algo") +
      coord_flip() + geom_hline(yintercept = 1) + geom_hline(yintercept = 0, linewidth = 0.2)
  )
}
```

```{r explore-units}
# ASV vs OTU vs OTU_derived, at the default settings (Q5).
res_comp_tax |>
  filter(dataset == first$dataset, metrics %in% c("F1_score", "TPR")) |>
  mutate(algo = sub("^(itsx|otu)_", "", algo)) |>
  filter_default_settings() |>
  filter(tax_level %in% c("Family", "Genus", "Gen_sp")) |>
  ggplot(aes(x = unit, y = values, fill = algo)) +
  geom_boxplot(alpha = 0.7) +
  facet_grid(metrics ~ factor(tax_level, levels = tax_order)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r explore-controls}
# The two control metrics the per-unit scoring adds (OD 7).
res_comp_tax |>
  filter(dataset == first$dataset, unit == "ASV",
         metrics %in% c("ext_fungi", "ext_correct", "NA_fake")) |>
  filter(tax_level %in% c("Kingdom", "Genus")) |>
  filter_default_settings() |>
  left_join(select(db_meta, db, simplification), by = "db") |>
  ggplot(aes(x = values, y = db, fill = algo)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_grid(metrics ~ tax_level) +
  xlim(c(0, 1))
```

```{r explore-tpr-tnr}
res_first |>
  filter(metrics %in% c("TPR", "TNR")) |>
  filter_default_settings() |>
  filter(!tax_level %in% c("Kingdom", "Phylum", "Class")) |>
  pivot_wider(names_from = metrics, values_from = values) |>
  left_join(select(db_meta, db, simplification), by = "db") |>
  ggplot(aes(x = TPR, y = TNR, pch = algo, color = simplification)) +
  geom_point(size = 3, alpha = 0.8) +
  facet_wrap(~ factor(tax_level, levels = tax_order))
```

```{r explore-tc-bar}
sintax_seed_db <- paste0("Family_", full_name_for("sintax", seed_taxonomy_db, 0.5))
tc_cols <- c(sintax_seed_db, paste0(c("Family_", "Genus_"), preference_pattern))

# psmelt() drops the all-NA rank columns, which happens with the mini_*
# databases: skip the views instead of failing in tc_bar().
empty_tc_cols <- tc_cols[vapply(tc_cols, \(col) all(is.na(tt_first[[col]])), logical(1))]
if (length(empty_tc_cols) > 0) {
  message("explore-tc-bar skipped, rank columns with no assignment: ",
          paste(empty_tc_cols, collapse = ", "))
} else {
  print(tc_bar(first$d_all_taxo,
               rank_1 = paste0("Family_", preference_pattern),
               rank_2 = sintax_seed_db,
               color_rank = paste0("Genus_", preference_pattern)))

  tc_circle(first$d_all_taxo,
            sintax_seed_db,
            paste0("Family_", preference_pattern))
}
```
