---
title: "02 — Q1: algorithm and parameters"
author: "Adrien Taudière"
---

Effect of the classification algorithm and its parameters on the mock
community. Figures: Q1.2 (per-rank × method), Q1.3 (bootstrap sweep), Q1.4
(blastn vote sweep); tables: Q1.6 (means and win matrix), Q1.7 (mixed model).

```{r setup}
here::i_am("analysis/02_q1_methods.qmd")  # _quarto.yml would otherwise make here() root = analysis/
source(here::here("analysis/00_setup.R"))
res_comp_tax <- cache_read("res_comp_tax") |> default_view()
```

## Q1.2 — Per-rank × method (ACC, F1, MCC)

Comparing the four methods at bootstrap 0.5; blastn shown with `rel_majority`
vote (the default; see Q1.4 for the vote-algorithm sweep).

```{r q1-2}
fig_q1_data <- res_comp_tax |>
  filter(metrics %in% c("ACC", "F1_score", "MCC")) |>
  filter_default_settings() |>
  mutate(tax_level = factor(tax_level, levels = tax_order))

fig_q1_methods <- fig_q1_data |>
  ggplot(aes(x = algo, y = values, fill = db)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_grid(metrics ~ tax_level) +
  labs(x = "Method", y = "Value", fill = "Database") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

save_fig(fig_q1_methods, "fig_q1_methods", width = 16, height = 7)
```

## Q1.3 — Bootstrap sweep (dada2 & sintax)

F1 and MCC across `min_bootstrap ∈ {0.4, 0.5, 0.6}` per rank.

```{r q1-3}
fig_q1_bootstrap_data <- res_comp_tax |>
  filter(metrics %in% c("F1_score", "MCC")) |>
  filter(algo %in% c("dada2", "sintax")) |>
  filter(!startsWith(db, "mini_")) |>
  mutate(tax_level = factor(tax_level, levels = tax_order))

fig_q1_bootstrap <- fig_q1_bootstrap_data |>
  ggplot(aes(x = factor(bootstrap_num), y = values, fill = db)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_grid(metrics + algo ~ tax_level) +
  labs(x = "Min bootstrap", y = "Value", fill = "Database") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

save_fig(fig_q1_bootstrap, "fig_q1_bootstrap", width = 16, height = 9)
```

## Q1.4 — Blastn vote-algorithm sweep

F1 and MCC across `c("rel_majority", "abs_majority", "unanimity")` per rank × DB,
at the default `min_id = 95` (the `min_id` and `lca_cutoff` sweeps get their own
figures in ROADMAP 0.8).

```{r q1-4}
fig_q1_blastn_data <- res_comp_tax |>
  filter(metrics %in% c("F1_score", "MCC")) |>
  filter(algo == "blastn", min_id == 95) |>
  filter(!startsWith(db, "mini_")) |>
  mutate(tax_level = factor(tax_level, levels = tax_order))

fig_q1_blastn <- fig_q1_blastn_data |>
  ggplot(aes(x = vote_algorithm, y = values, fill = db)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_grid(metrics ~ tax_level) +
  labs(x = "Vote algorithm", y = "Value", fill = "Database") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

save_fig(fig_q1_blastn, "fig_q1_blastn_vote", width = 16, height = 7)
```

## Q1.6 — Per-rank method comparison table and win matrix

```{r q1-6-table}
method_summary_tbl <- res_comp_tax |>
  filter(metrics %in% c("F1_score", "ACC")) |>
  filter_default_settings() |>
  group_by(tax_level, algo, metrics) |>
  summarise(
    mean_val = round(mean(values, na.rm = TRUE), 3),
    sd_val   = round(sd(values, na.rm = TRUE), 3),
    .groups  = "drop"
  ) |>
  mutate(mean_sd = paste0(mean_val, " (±", sd_val, ")")) |>
  pivot_wider(
    id_cols     = c(tax_level, algo),
    names_from  = metrics,
    values_from = mean_sd
  ) |>
  mutate(tax_level = factor(tax_level, levels = tax_order)) |>
  arrange(tax_level, algo)

knitr::kable(
  method_summary_tbl,
  col.names = c("Rank", "Method", "ACC", "F1"),
  caption   = "Mean (±sd) F1 and ACC per method × rank (bootstrap = 0.5, averaged across DBs)"
)
```

```{r q1-6-wins}
win_matrix_tbl <- res_comp_tax |>
  filter(metrics == "F1_score") |>
  filter_default_settings() |>
  group_by(tax_level, db) |>
  slice_max(values, n = 1, with_ties = FALSE) |>
  ungroup() |>
  count(tax_level, algo, name = "n_wins") |>
  pivot_wider(
    names_from  = algo,
    values_from = n_wins,
    values_fill = 0L
  ) |>
  mutate(tax_level = factor(tax_level, levels = tax_order)) |>
  arrange(tax_level)

knitr::kable(
  win_matrix_tbl,
  caption = "Win matrix: number of DBs where each method achieves highest F1 per rank (bootstrap = 0.5)"
)
```

## Q1.7 — Statistical test: F1 ~ method + (1|db) per rank

Linear mixed model with DB as random intercept, testing whether the method
fixed effect explains significant variance in F1 (LRT vs. null model). Requires
`lme4`.

```{r q1-7}
lmm_res_list <- lapply(
  intersect(tax_order, unique(res_comp_tax$tax_level)),
  function(rank) {
    dat <- res_comp_tax |>
      filter(metrics == "F1_score", tax_level == rank) |>
      filter_default_settings()
    if (nrow(dat) < 8 || dplyr::n_distinct(dat$db) < 3) {
      return(NULL)
    }
    tryCatch({
      m_full <- lme4::lmer(values ~ algo + (1 | db), data = dat, REML = FALSE)
      m_null <- lme4::lmer(values ~ 1 + (1 | db), data = dat, REML = FALSE)
      lrt    <- anova(m_null, m_full)
      tibble::tibble(
        tax_level = rank,
        LRT_chisq = round(lrt$Chisq[2], 3),
        df        = as.integer(lrt[["Chi Df"]][2]),
        p_value   = round(lrt[["Pr(>Chisq)"]][2], 4),
        signif    = dplyr::case_when(
          lrt[["Pr(>Chisq)"]][2] < 0.001 ~ "***",
          lrt[["Pr(>Chisq)"]][2] < 0.01  ~ "**",
          lrt[["Pr(>Chisq)"]][2] < 0.05  ~ "*",
          TRUE ~ ""
        )
      )
    }, error = function(e) {
      tibble::tibble(tax_level = rank, LRT_chisq = NA_real_,
                     df = NA_integer_, p_value = NA_real_, signif = "ERROR")
    })
  }
)

lmm_results <- dplyr::bind_rows(lmm_res_list) |>
  mutate(tax_level = factor(tax_level, levels = tax_order)) |>
  arrange(tax_level)

knitr::kable(
  lmm_results,
  caption = "LRT: F1 ~ method + (1|db) vs. null (1|db) per rank. *** p<0.001, ** p<0.01, * p<0.05."
)
```
