---
title: "03 — Q2: reference database and simplification"
author: "Adrien Taudière"
---

Effect of the reference database and of its simplification (Fungi-only,
primer-trimmed `_cut`). Figures: Q2.1 (per-DB accuracy), Q2.2 (paired
simplification effect), Q2.5 (DB size vs compute cost and accuracy).

```{r setup}
here::i_am("analysis/03_q2_databases.qmd")  # _quarto.yml would otherwise make here() root = analysis/
source(here::here("analysis/00_setup.R"))
res_comp_tax <- cache_read("res_comp_tax") |> default_view()
benchmark_costs <- tar_read(benchmark_costs, store = store_assign_taxo)

# benchmark_costs joins res_comp_tax on the assignment names. A store built
# before the 2026-09-17 grid holds other names, and the join would come back
# empty without any error: stop instead.
matched <- sum(benchmark_costs$phase %in% res_comp_tax$method_db)
if (matched == 0) {
  stop(
    "No assignment of benchmark_costs matches res_comp_tax: ",
    basename(store_assign_taxo),
    " predates the current grid. Rerun the assign_taxo project."
  )
}
```

## Q2.1 — Per-DB accuracy

Default settings (bootstrap 0.5 / rel_majority). `db_meta` (which
simplification step each DB represents) comes from `config.R`.

```{r q2-1}
fig_q2_base <- res_comp_tax |>
  filter(metrics %in% c("F1_score", "MCC")) |>
  filter_default_settings() |>
  left_join(db_meta, by = "db") |>
  mutate(
    tax_level      = factor(tax_level,      levels = tax_order),
    simplification = factor(simplification, levels = c("full", "Fungi", "cut", "Fungi+cut"))
  )

fig_q2_db <- fig_q2_base |>
  ggplot(aes(x = db, y = values, fill = simplification)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_grid(metrics + algo ~ tax_level) +
  labs(x = "Database", y = "Value", fill = "Simplification") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1))

save_fig(fig_q2_db, "fig_q2_db", width = 18, height = 14)
```

## Q2.2 — Simplification effect (paired Δ F1/MCC)

Paired comparisons within each source release: full → Fungi and
Fungi → Fungi+cut, derived from `db_meta` in `config.R`, so they follow the
databases actually benchmarked.

```{r q2-2-table}
# Signed delta (simplified − full) for one pair of databases.
db_pair_delta <- function(data, db_full, db_simp, label) {
  data |>
    filter(db %in% c(db_full, db_simp)) |>
    select(algo, tax_level, metrics, db, values) |>
    pivot_wider(names_from = db, values_from = values) |>
    mutate(
      delta      = .data[[db_simp]] - .data[[db_full]],
      comparison = label
    ) |>
    select(algo, tax_level, metrics, comparison, delta)
}

# Pairs (from -> to) of simplification steps within the same source.
pair_steps <- function(from, to) {
  inner_join(
    filter(db_meta, simplification == from),
    filter(db_meta, simplification == to),
    by = "db_base",
    suffix = c("_full", "_simp")
  ) |>
    transmute(db_full, db_simp, label = paste0(db_base, ": ", from, " → ", to))
}

db_pairs <- bind_rows(pair_steps("full", "Fungi"), pair_steps("Fungi", "Fungi+cut"))

tab_q2_simplif <- purrr::pmap(
  db_pairs,
  \(db_full, db_simp, label) db_pair_delta(fig_q2_base, db_full, db_simp, label)
) |>
  bind_rows() |>
  mutate(tax_level = factor(tax_level, levels = tax_order))

tab_q2_simplif |>
  group_by(comparison, algo, metrics, tax_level) |>
  summarise(delta_mean = round(mean(delta, na.rm = TRUE), 3), .groups = "drop") |>
  pivot_wider(names_from = tax_level, values_from = delta_mean) |>
  arrange(comparison, metrics, algo) |>
  knitr::kable(caption = "Δ F1/MCC (simplified − full). Positive = simplification helps.")
```

```{r q2-2-fig}
fig_q2_simplif <- tab_q2_simplif |>
  ggplot(aes(x = algo, y = delta, fill = comparison)) +
  geom_hline(yintercept = 0, linewidth = 0.4, linetype = "dashed") +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_grid(metrics ~ tax_level) +
  labs(x = "Method", y = "Δ (simplified − full)", fill = "Comparison") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

save_fig(fig_q2_simplif, "fig_q2_simplification", width = 18, height = 8)
```

## Q2.5 — DB size vs compute cost and accuracy

Does a smaller (filtered / trimmed) DB pay for itself in runtime without
losing F1? Uses `benchmark_costs` (see `05_m2_costs.qmd` for its provenance).

```{r q2-5}
db_size_tbl <- benchmark_costs |>
  filter(!startsWith(db, "mini_")) |>
  mutate(db_size_mb = file.size(here(db_path)) / 1e6) |>
  group_by(db, db_size_mb) |>
  summarise(mean_wall_h = mean(wall_time_s, na.rm = TRUE) / 3600,
            .groups = "drop")

f1_by_db <- res_comp_tax |>
  filter(metrics == "F1_score", tax_level == "Gen_sp") |>
  filter_default_settings() |>
  group_by(db) |>
  summarise(mean_f1 = mean(values, na.rm = TRUE), .groups = "drop")

q2_5_data <- left_join(db_size_tbl, f1_by_db, by = "db") |>
  left_join(db_meta |> select(db, db_base, simplification), by = "db")

p_size_time <- q2_5_data |>
  ggplot(aes(x = db_size_mb, y = mean_wall_h, color = db_base, label = simplification)) +
  geom_point(size = 3) +
  ggrepel::geom_label_repel(size = 2.5, show.legend = FALSE) +
  labs(x = "DB size (MB)", y = "Mean wall time (h)", color = "DB family") +
  theme_bw()

p_size_f1 <- q2_5_data |>
  ggplot(aes(x = db_size_mb, y = mean_f1, color = db_base, label = simplification)) +
  geom_point(size = 3) +
  ggrepel::geom_label_repel(size = 2.5, show.legend = FALSE) +
  labs(x = "DB size (MB)", y = "Mean F1 (Gen_sp)", color = "DB family") +
  theme_bw()

fig_q2_5 <- p_size_time / p_size_f1 + patchwork::plot_layout(guides = "collect")
save_fig(fig_q2_5, "fig_q2_5_size_cost", width = 8, height = 10)
```
