---
title: "05 — M2: compute cost and carbon footprint"
author: "Adrien Taudière"
---

Accuracy versus compute cost of each assignment target. `benchmark_costs`
comes from the per-target {autometric} logs written inside the crew workers
(`R/autometric_helpers.R`, ROADMAP S1.1): wall time, peak resident memory and
mean CPU of the newest run of each target. M2.2 converts wall time to CO₂eq
with `greenAlgoR`.

Until the production rerun that follows S1.1, this chapter reads a
`benchmark_costs` object computed from the legacy 2025-02 log; treat the
figures as placeholders (ROADMAP, M2 re-validation).

```{r setup}
here::i_am("analysis/05_m2_costs.qmd")  # _quarto.yml would otherwise make here() root = analysis/
source(here::here("analysis/00_setup.R"))
res_comp_tax    <- cache_read("res_comp_tax")
benchmark_costs <- tar_read(benchmark_costs, store = store_assign_taxo)
```

## M2.1 — Accuracy vs cost (F1 × wall time)

The manuscript's practical recommendation: which method gives the best
accuracy per unit of compute? Default settings only.

```{r m2-1}
f1_all_ranks <- res_comp_tax |>
  filter(metrics == "F1_score") |>
  filter_default_settings() |>
  mutate(tax_level = factor(tax_level, levels = tax_order))

accuracy_vs_cost <- f1_all_ranks |>
  left_join(
    benchmark_costs |> select(phase, wall_time_s, peak_resident_mb, mean_cpu_pct),
    by = c("method_db" = "phase")
  )

fig_m2_cost_species <- accuracy_vs_cost |>
  filter(tax_level == "Gen_sp", !is.na(wall_time_s)) |>
  ggplot(aes(x = wall_time_s / 3600, y = values, color = algo)) +
  geom_point(size = 3, alpha = 0.85) +
  ggrepel::geom_label_repel(aes(label = db), size = 2.2, max.overlaps = 20,
                            show.legend = FALSE) +
  labs(x = "Wall time (hours)", y = "F1 score (Gen_sp rank)", color = "Method") +
  theme_bw()

save_fig(fig_m2_cost_species, "fig_m2_cost_species", width = 8, height = 6)
```

```{r m2-1-allranks}
fig_m2_cost_allranks <- accuracy_vs_cost |>
  filter(!is.na(wall_time_s)) |>
  ggplot(aes(x = wall_time_s / 3600, y = values, color = algo, shape = db)) +
  geom_point(size = 2.5, alpha = 0.8) +
  facet_wrap(~tax_level, nrow = 1) +
  labs(x = "Wall time (hours)", y = "F1 score", color = "Method", shape = "Database") +
  theme_bw() +
  theme(legend.position = "bottom")

save_fig(fig_m2_cost_allranks, "fig_m2_cost_allranks", width = 18, height = 5)
```

## M2.2 — CO₂eq per assignment target

Carbon footprint (gCO2eq) estimated via `greenAlgoR::ga_footprint()` from
`wall_time_s`. Hardware parameters default to the machine's detected RAM and
WORLD carbon intensity; `n_cores` is `n_threads` from `config.R`.

```{r m2-2-co2}
benchmark_costs <- benchmark_costs |>
  rowwise() |>
  mutate(
    co2eq_gCO2 = greenAlgoR::ga_footprint(
      runtime_h      = wall_time_s / 3600,
      n_cores        = n_threads,
      add_ref_values = FALSE
    )$carbon_footprint_total_gCO2
  ) |>
  ungroup()

accuracy_vs_co2 <- f1_all_ranks |>
  left_join(
    benchmark_costs |> select(phase, co2eq_gCO2),
    by = c("method_db" = "phase")
  )

fig_m2_co2_species <- accuracy_vs_co2 |>
  filter(tax_level == "Gen_sp", !is.na(co2eq_gCO2)) |>
  ggplot(aes(x = co2eq_gCO2, y = values, color = algo)) +
  geom_point(size = 3, alpha = 0.85) +
  ggrepel::geom_label_repel(
    aes(label = db), size = 2.2, max.overlaps = 20, show.legend = FALSE
  ) +
  labs(
    x = expression(CO[2]*"eq (gCO2)"),
    y = "F1 score (Gen_sp rank)",
    color = "Method"
  ) +
  theme_bw()

save_fig(fig_m2_co2_species, "fig_m2_co2_species", width = 8, height = 6)
```

```{r m2-2-allranks}
fig_m2_co2_allranks <- accuracy_vs_co2 |>
  filter(!is.na(co2eq_gCO2)) |>
  ggplot(aes(x = co2eq_gCO2, y = values, color = algo, shape = db)) +
  geom_point(size = 2.5, alpha = 0.8) +
  facet_wrap(~tax_level, nrow = 1) +
  labs(
    x = expression(CO[2]*"eq (gCO2)"),
    y = "F1 score",
    color = "Method",
    shape = "Database"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

save_fig(fig_m2_co2_allranks, "fig_m2_co2_allranks", width = 18, height = 5)
```

## M2.3 — DADA2 preprocessing cost

Cost of the denoising step, attributable separately from the assignments
(`benchmark_costs_dada2` in `store_dada2`).

```{r m2-3}
tar_read(benchmark_costs_dada2, store = here("store_dada2")) |>
  arrange(desc(wall_time_s)) |>
  knitr::kable(
    digits = 1,
    caption = "DADA2 phases: wall time (s), peak resident memory (MB), mean CPU (%)"
  )
```
