---
title: "06 — D1a: cross-validation"
author: "Adrien Taudière"
---

k-fold cross-validation of each (method, database) pair on the database itself
(`pipelines/cross_val.R` → `store_cross_val`). `remove_tested = TRUE` is the
standard estimate (held-out sequences removed from the reference);
`remove_tested = FALSE` is the leaked upper bound. Figures parallel to Q1 and
Q2 are still to be finalised (ROADMAP, D1a).

```{r setup}
here::i_am("analysis/06_cross_validation.qmd")  # _quarto.yml would otherwise make here() root = analysis/
source(here::here("analysis/00_setup.R"))
cv_res <- tar_read(cv_results, store = store_cross_val) |>
  filter(!startsWith(db, "mini_")) |>
  mutate(
    tax_level = factor(tax_level, levels = intersect(tax_order, unique(tax_level))),
    variant   = ifelse(remove_tested, "standard", "leaked")
  )
```

## Good classifications per rank, method and database

```{r cv-good}
fig_cv_good <- cv_res |>
  filter(metric == "good_classifications") |>
  ggplot(aes(x = mean, y = factor(method), fill = db)) +
  geom_col(position = position_dodge()) +
  geom_errorbar(aes(xmin = mean - sd, xmax = mean + sd),
                position = position_dodge(width = 0.9), width = 0.3) +
  facet_grid(variant ~ tax_level) +
  labs(x = "Proportion of correct assignments (mean ± sd over folds)",
       y = "Method", fill = "Database") +
  theme_bw()

save_fig(fig_cv_good, "fig_cv_good_classifications", width = 16, height = 7)
```

## Standard vs leaked (database leakage effect)

```{r cv-leak}
cv_res |>
  filter(metric == "good_classifications") |>
  select(method, db, tax_level, variant, mean) |>
  pivot_wider(names_from = variant, values_from = mean) |>
  mutate(leak_gain = leaked - standard) |>
  group_by(method, tax_level) |>
  summarise(mean_leak_gain = round(mean(leak_gain, na.rm = TRUE), 3), .groups = "drop") |>
  pivot_wider(names_from = tax_level, values_from = mean_leak_gain) |>
  knitr::kable(caption = "Gain in correct assignments when tested sequences stay in the reference (leaked − standard), averaged over databases.")
```

## All metrics

```{r cv-all}
cv_res |>
  ggplot(aes(x = mean, y = factor(method), fill = db)) +
  geom_col(position = position_dodge()) +
  facet_grid(metric + variant ~ tax_level, scales = "free_x") +
  labs(x = "Mean over folds", y = "Method", fill = "Database") +
  theme_bw()
```
