---
title: "Benchmark analysis"
author: "Adrien Taudière"
---


```{r}
library("conflicted")
library("MiscMetabar")
library("targets")
library("tarchetypes")
library("here")
library("tibble")
library("tidyr")
library("autometric")

conflicted::conflicts_prefer(dplyr::filter)
```


```{r}
here::i_am("analysis/benchmark.qmd")
devtools::load_all("/home/adrien/Nextcloud/IdEst/Projets/pqverse/pqverse_pkg/MiscMetabar/")
devtools::load_all("/home/adrien/Nextcloud/IdEst/Projets/pqverse/pqverse_pkg/dbpq/")
devtools::load_all("/home/adrien/Nextcloud/IdEst/Projets/pqverse/pqverse_pkg/greenAlgoR/")
source("/home/adrien/Nextcloud/IdEst/Projets/pqverse/pqverse_pkg/comparpq/R/compare_taxo.R")
source("/home/adrien/Nextcloud/IdEst/Projets/pqverse/pqverse_pkg/comparpq/R/taxtab_modification.R")
source("/home/adrien/Nextcloud/IdEst/Projets/pqverse/pqverse_pkg/comparpq/R/compare_taxo_plot.R")
```

# Import value from store_assign_taxo

```{r}
here::i_am("analysis/benchmark.qmd")
source(here("config.R"))
source(here("R/values_map_for_qmd.R"))
```

Load the targets store and phyloseq object. The `store = "store_assign_taxo"` argument
resolves to the on-disk store alongside the pipeline script.

```{r, results = "hide"}
# Ensure we are in the project root so the store path resolves correctly.
setwd(here::here())
d_all_taxo <- tar_read(d_all_taxo, store = "store_assign_taxo")
taxo_mock  <- tar_read(taxo_mock,  store = "store_assign_taxo") |>
  rename(Gen_sp = Species)
```

```{r}
d_all_taxo <- d_all_taxo |>
  taxtab_replace_pattern_by_NA(
    patterns = ".*_sp$",
    progress_bar = TRUE,
    taxonomic_ranks =
      colnames(d_all_taxo@tax_table)[grep("^Species_", colnames(d_all_taxo@tax_table))]
  ) |>
  taxtab_replace_pattern_by_NA(
    patterns = c(
      ".*_incertae_sedis",
      ".*_Incertae_sedis",
      "unclassified.*",
      "^$"
    ),
    progress_bar = TRUE
  ) |>
  simplify_taxo()
```

```{r}
taxtab_d_all <- d_all_taxo@tax_table |> 
  as.data.frame() 

cl_names <- gsub("Kingdom|Phylum|Phyla|Class|Order|Family|Genus|Species", NA, unique(gsub("Kingdom_|Phyla_|Phylum_|Class_|Order_|Family_|Genus_|Species_", "",
            colnames(d_all_taxo@tax_table)))) |>
  na.omit()  |>
  as.vector()

cl_names <- grep("Taxa_name_", cl_names, invert = TRUE,value=TRUE)

for(cl in cl_names) {
  if(paste0("Species_", cl)%in%colnames(taxtab_d_all)) {
      sp <- as.vector(taxtab_d_all[,paste0("Species_", cl)])
  } else {
    sp <- NA
  }
  if(paste0("Genus_", cl)%in%colnames(taxtab_d_all)) {
      gen <- as.vector(taxtab_d_all[,paste0("Genus_", cl)])
  } else {
    gen <- NA
  }
  
  if(sum(grepl("_", sp)) == sum(!is.na(sp))) {
    taxtab_d_all[paste0("Gen_sp_", cl)] <- sp
  } else  {
    taxtab_d_all[paste0("Gen_sp_", cl)] <- paste(gen, sp, sep = "_")
  }
   taxtab_d_all[grepl("NA", as.vector(taxtab_d_all[paste0("Gen_sp_", cl)][,1])), paste0("Gen_sp_", cl)] <- NA
}

d_all_taxo@tax_table <- tax_table(as.matrix(taxtab_d_all))
```

## Create column using consensus with clear majority 

```{r}
d_all_taxo <-  d_all_taxo |>
  resolve_taxo_conflict(
    pattern_tax_ranks = c("Kingdom","Phylum", "Class", "Order", "Family", "Genus", "Gen_sp"),
    method = "rel_majority",
    new_names = paste0(c("Kingdom","Phylum", "Class", "Order", "Family", "Genus", "Gen_sp"), "_", "rel_majority_consensus"),
    replace_collapsed_rank_by_NA = TRUE
  )  |>
  resolve_taxo_conflict(
    pattern_tax_ranks = c("Kingdom","Phylum", "Class", "Order", "Family", "Genus", "Gen_sp"),
    method = "preference",
        new_names = paste0(c("Kingdom","Phylum", "Class", "Order", "Family", "Genus", "Gen_sp"), "_", "preference_consensus"),
    preference_pattern = "sintax__EUK_ITS_v1_9_3___0.5",  # replace by sintax__mini_EUK_ITS_v1_9_3___0.5 if mini db are used 
    replace_collapsed_rank_by_NA = TRUE
  )  |>
  resolve_taxo_conflict(
    pattern_tax_ranks = c("Kingdom","Phylum", "Class", "Order", "Family", "Genus", "Gen_sp"),
    method = "unanimity",
    new_names = paste0(c("Kingdom","Phylum", "Class", "Order", "Family", "Genus", "Gen_sp"), "_", "unanimity_consensus"),
    replace_collapsed_rank_by_NA = TRUE
  )
```

```{r}
# Q3.1: add consensus + abs_majority (strict=FALSE) and strict=TRUE variants of
# all 5 methods. Q3.3: rel_majority with nb_agree_threshold = 2 and 3.
.ranks_base    <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Gen_sp")
.pref_pattern  <- "sintax__EUK_ITS_v1_9_3___0.5"

extra_consensus_configs <- list(
  list(method = "consensus",    strict = FALSE, nb = 1L, suffix = "consensus_consensus"),
  list(method = "abs_majority", strict = FALSE, nb = 1L, suffix = "abs_majority_consensus"),
  list(method = "rel_majority", strict = TRUE,  nb = 1L, suffix = "rel_majority_strict_consensus"),
  list(method = "preference",   strict = TRUE,  nb = 1L, suffix = "preference_strict_consensus"),
  list(method = "unanimity",    strict = TRUE,  nb = 1L, suffix = "unanimity_strict_consensus"),
  list(method = "consensus",    strict = TRUE,  nb = 1L, suffix = "consensus_strict_consensus"),
  list(method = "abs_majority", strict = TRUE,  nb = 1L, suffix = "abs_majority_strict_consensus"),
  list(method = "rel_majority", strict = FALSE, nb = 2L, suffix = "rel_majority_nb2_consensus"),
  list(method = "rel_majority", strict = FALSE, nb = 3L, suffix = "rel_majority_nb3_consensus")
)

for (cfg in extra_consensus_configs) {
  args <- list(
    physeq              = d_all_taxo,
    pattern_tax_ranks   = .ranks_base,
    method              = cfg$method,
    strict              = cfg$strict,
    nb_agree_threshold  = cfg$nb,
    new_names           = paste0(.ranks_base, "_", cfg$suffix),
    replace_collapsed_rank_by_NA = TRUE
  )
  if (cfg$method == "preference") {
    args$preference_pattern <- .pref_pattern
  }
  d_all_taxo <- do.call(resolve_taxo_conflict, args)
}
```


# Compute performance metrics

## Proportion of NA

```{r}
df_na <- data.frame("method_db"=names(colSums(is.na(d_all_taxo@tax_table)/ntaxa(d_all_taxo))),
           "prop_na" = as.vector(colSums(is.na(d_all_taxo@tax_table))/ntaxa(d_all_taxo))) |>
  mutate(tax_level = gsub("_.*", "", method_db)) |>
  mutate(tax_level = substr(tax_level, 1, 1)) |>
  filter(tax_level%in%c("K", "P", "C", "O", "F", "G", "S")) |>
  mutate(bootstrap = gsub(".*__", "", method_db)) |>
  mutate(db = sub(".*__", "", gsub("___.*", "", method_db))) |>
  mutate(algo = method_db) |>
  mutate(algo = ifelse(grepl("dada2", algo), "dada2", algo)) |>
  mutate(algo = ifelse(grepl("sintax", algo), "sintax", algo)) |>
  mutate(algo = ifelse(grepl("lca", algo), "lca", algo)) |>
  mutate(algo = ifelse(grepl("blastn", algo), "blastn", algo)) |>
  mutate(algo = ifelse(grepl("idtaxa", algo), "idtaxa", algo)) |>
  mutate(algo = ifelse(grepl("rel_majority_consensus", algo), "rel_majority_consensus", algo)) |>
  mutate(algo = ifelse(grepl("preference_consensus", algo), "preference_consensus", algo)) |>
  mutate(algo = ifelse(grepl("unanimity_consensus", algo), "unanimity_consensus", algo)) |>
  mutate(db = ifelse(grepl("unanimity_consensus", algo), NA, db)) |>
  mutate(db = ifelse(grepl("preference_consensus", algo), NA, db)) |>
  mutate(db = ifelse(grepl("rel_majority_consensus", algo), NA, db)) |>
  mutate(bootstrap = ifelse(grepl("unanimity_consensus", algo), NA, bootstrap)) |>
  mutate(bootstrap = ifelse(grepl("rel_majority_consensus", algo), NA, bootstrap)) |>
  mutate(bootstrap = ifelse(grepl("preference", algo), NA, bootstrap)) 

df_na |>
  dplyr::filter(algo%in%c("unanimity_consensus", "preference_consensus","rel_majority_consensus")) |> # "idtaxa"
  ggplot() +
  geom_bar(aes(x=prop_na, y=db, fill=db),stat = "identity", position = position_dodge()) +
  geom_vline(xintercept = sum(grepl("external|fake",taxa_names(d_all_taxo)))/ntaxa(d_all_taxo)) +
facet_grid(algo~factor(tax_level, levels=c("K", "P", "C", "O", "F", "G", "S"))) +
  xlim(c(0,1)) + no_legend()

df_na  |>
  dplyr::filter(algo%in%c("dada2", "sintax", "lca", "blastn")) |> # "idtaxa"
  ggplot() +
  geom_bar(aes(x=prop_na, y=db, fill=db),stat = "identity", position = position_dodge()) +
  geom_vline(xintercept = sum(grepl("external|fake",taxa_names(d_all_taxo)))/ntaxa(d_all_taxo)) +
facet_grid(algo~factor(tax_level, levels=c("K", "P", "C", "O", "F", "G", "S"))) +
  xlim(c(0,1)) + no_legend()
```



## Create a dataframe of taxonomic level of interest

```{r}
ranks_df_list <- list()

fullnames_vec <- values_map |>
  pull(full_name)
  

all_consensus_suffixes <- c(
  "rel_majority_consensus",
  "preference_consensus",
  "unanimity_consensus",
  "consensus_consensus",
  "abs_majority_consensus",
  "rel_majority_strict_consensus",
  "preference_strict_consensus",
  "unanimity_strict_consensus",
  "consensus_strict_consensus",
  "abs_majority_strict_consensus",
  "rel_majority_nb2_consensus",
  "rel_majority_nb3_consensus"
)

for (db_name in c(all_consensus_suffixes, fullnames_vec)){
  ranks_df_list[[db_name]] <-   c(
    paste0(c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Gen_sp"),
           paste0("_", db_name))
  )
}

ranks_df <- do.call(data.frame, ranks_df_list)

if(sum(!as.vector(unlist(ranks_df)) %in% colnames(d_all_taxo@tax_table)) != 0) {
  as.vector(unlist(ranks_df))[!as.vector(unlist(ranks_df)) %in% colnames(d_all_taxo@tax_table)]
  stop("there is a big pb!")
}
```


```{r}
res_comp_tax <- tc_metrics_mock(d_all_taxo,
                                ranks_df = ranks_df,
                                true_values_df = taxo_mock)

res_comp_tax <- res_comp_tax |>
  mutate(bootstrap = gsub(".*__", "", method_db)) |>
  mutate(algo = gsub("__.*", "", method_db)) |>
  mutate(db = gsub(".*__", "", gsub("___.*", "", method_db)))
```

```{r}
# M1.1: enrich res_comp_tax with vote_algorithm, bootstrap_num, and
# TAR/TDR aliases (TAR = PPV = TP/(TP+FP); TDR = TPR = TP/(TP+FN)).
res_comp_tax <- res_comp_tax |>
  mutate(
    vote_algorithm = ifelse(
      algo == "blastn",
      stringr::str_extract(method_db, "(?<=\\.\\.\\.)[a-z_]+(?=\\.\\.\\.)"),
      NA_character_
    ),
    bootstrap_num = suppressWarnings(
      as.numeric(stringr::str_extract(bootstrap, "^[0-9.]+"))
    )
  )

res_comp_tax <- dplyr::bind_rows(
  res_comp_tax,
  res_comp_tax |> dplyr::filter(metrics == "PPV") |> dplyr::mutate(metrics = "TAR"),
  res_comp_tax |> dplyr::filter(metrics == "TPR") |> dplyr::mutate(metrics = "TDR")
)

# Consensus-row metadata: is_consensus, strict flag, nb_agree_threshold value.
res_comp_tax <- res_comp_tax |>
  dplyr::mutate(
    is_consensus   = !algo %in% c("dada2", "sintax", "lca", "blastn"),
    consensus_strict = is_consensus & grepl("_strict_", method_db),
    nb_agree       = dplyr::case_when(
      grepl("_nb2_", method_db) ~ 2L,
      grepl("_nb3_", method_db) ~ 3L,
      TRUE ~ 1L
    )
  )
```


```{r}
plot_tc_metrics_mock <- function(tc_metrics_mock,
                                 metric=NULL,
                                 modality_x=NULL,
                                 modality_line=NULL,
                                modality_column=NULL,
                                modality_fill=NULL,
                                taxonomic_ranks=NULL,
                                legend= TRUE){

  if(is.null(taxonomic_ranks)){
    taxonomic_ranks <- unique(tc_metrics_mock$tax_level)
  } else {
    tc_metrics_mock <- tc_metrics_mock |>
      dplyr::filter(tax_level %in% taxonomic_ranks)
  }

  if(is.null(modality_fill)){
    modality_fill <- "method_db"
  }

  if(is.null(modality_x)){
    modality_x <- "tax_level"
  }

  tc_metrics_mock$tax_level <- factor(tc_metrics_mock$tax_level, taxonomic_ranks)

  if(!is.null(metric)){
    tc_metrics_mock <- dplyr::filter(tc_metrics_mock, metrics == metric)
  } else {
if(modality_fill != "metrics" &&
   modality_column != "metrics" &&
   modality_line != "metrics" &&
   modality_x != "metrics"
) {
  stop("You must either choose one metric (parameter metric) or choose the
       value 'metrics' in one of parameters modality_x, modality_line,
       modality_column or modality_fill")
}
  }

  if(is.null(modality_column) && is.null(modality_line)) {
    p <- tc_metrics_mock |>
      ggplot(aes(fill=.data[[modality_fill]], y=values, x=.data[[modality_x]])) +
      geom_bar(stat = "identity", position=position_dodge())
  } else if(is.null(modality_column) && !is.null(modality_line)) {
    p <- tc_metrics_mock |>
      ggplot(aes(fill=.data[[modality_fill]], y=values, x=.data[[modality_x]])) +
      geom_bar(stat = "identity", position=position_dodge()) +
      facet_grid(.data[[modality_line]]~., scales="free_y")
  } else if(!is.null(modality_column) && is.null(modality_line)) {
    p <- tc_metrics_mock |>
      ggplot(aes(fill=.data[[modality_fill]], y=values, x=.data[[modality_x]])) +
      geom_bar(stat = "identity", position=position_dodge()) +
      facet_grid(.~.data[[modality_column]], scales="free_x")
  } else if(!is.null(modality_column) && !is.null(modality_line)) {
    p <- tc_metrics_mock |>
      ggplot(aes(fill=.data[[modality_fill]], y=values, x=.data[[modality_x]])) +
      geom_bar(stat = "identity", position=position_dodge()) +
      facet_grid(vars(.data[[modality_line]]), vars(.data[[modality_column]]), scales="free")
  }

    if(!legend) {
      p <- p + no_legend()
    }

    return(p)
}
```


```{r}
plot_tc_metrics_mock(res_comp_tax, metric="ACC", modality_x ="tax_level") + no_legend()
```


```{r}
plot_tc_metrics_mock(res_comp_tax, modality_x ="algo", modality_line="metrics", modality_column = "tax_level", modality_fill = "db")  
```

```{r}
plot_tc_metrics_mock(res_comp_tax, metric="FP", modality_x ="db", modality_line="algo", modality_column = "tax_level", modality_fill = "bootstrap") + coord_flip()  + geom_hline(yintercept = 1)  + geom_hline(yintercept = 0, linewidth=0.2)
```

```{r}
plot_tc_metrics_mock(res_comp_tax, metric="MCC", modality_x ="db", modality_line="algo", modality_column = "tax_level", modality_fill = "bootstrap") + coord_flip()  + geom_hline(yintercept = 1)  + geom_hline(yintercept = 0, linewidth=0.2)
```

```{r}
plot_tc_metrics_mock(res_comp_tax, metric="F1_score", modality_x ="db", modality_line="algo", modality_column = "tax_level", modality_fill = "bootstrap") + coord_flip() + geom_hline(yintercept = 1)
```

```{r}
plot_tc_metrics_mock(res_comp_tax, metric="ACC", modality_x ="db", modality_line="algo", modality_column = "tax_level", modality_fill = "bootstrap") + coord_flip() + geom_hline(yintercept = 1)
```


```{r}
res_comp_tax |>
  dplyr::filter(metrics%in%c("ACC", "F1_score", "MCC")) |>
    dplyr::filter(bootstrap == 0.5 | grepl("consensus", algo) | algo=="blastn")  |>
  dplyr::filter(tax_level %in% c("Class", "Order", "Family","Genus", "Gen_sp")) |>
plot_tc_metrics_mock(modality_x ="tax_level", modality_line="metrics", modality_column = "algo", modality_fill = "db")

res_comp_tax |>
  dplyr::filter(metrics%in%c("ACC", "F1_score", "MCC", "FP", "FN")) |>
  dplyr::filter(tax_level %in% c("Class", "Order", "Family","Genus", "Gen_sp")) |>
    dplyr::filter(db=="Unite") |>
plot_tc_metrics_mock(modality_x ="bootstrap", modality_line="metrics", modality_column = "tax_level", modality_fill = "algo") +
  ggtitle("Unite")


res_comp_tax |>
  dplyr::filter(metrics%in%c("ACC", "F1_score", "MCC", "FP", "FN")) |>
  dplyr::filter(tax_level %in% c("Class", "Order", "Family","Genus", "Gen_sp")) |>
    dplyr::filter(db=="Unite_Fungi") |>
plot_tc_metrics_mock(modality_x ="bootstrap", modality_line="metrics", modality_column = "tax_level", modality_fill = "algo") +
  ggtitle("Unite fungi")

res_comp_tax |>
  dplyr::filter(metrics%in%c("ACC", "F1_score", "MCC", "FP", "FN")) |>
  dplyr::filter(tax_level %in% c("Class", "Order", "Family","Genus", "Gen_sp")) |>
    dplyr::filter(db=="EUK_ITS_v1_9_3") |>
plot_tc_metrics_mock(modality_x ="bootstrap", modality_line="metrics", modality_column = "tax_level", modality_fill = "algo") +
  ggtitle("EUK_ITS_v1_9_3")
```

```{r}
res_comp_tax |>
  dplyr::filter(metrics %in% c("TPR", "TNR")) |>
  mutate(db_unite = ifelse(grepl("Unite",method_db), "Unite", "Euk")) |>
  dplyr::filter(bootstrap == 0.5) |>
    dplyr::filter(!tax_level %in% c("Kingdom", "Phylum", "Class", "Gen_sp")) |>
    pivot_wider(names_from = metrics, values_from = values) |>
  ggplot(aes(x = TPR, y = TNR, pch=algo, color=db)) +
  geom_point(size=3, alpha=0.8)  +
  xlim(c(0.4,1)) + ylim(c(0.4,1)) +
  facet_wrap(db_unite~tax_level)
```

```{r}
p <- res_comp_tax |>
  dplyr::filter(metrics %in% c("TPR", "TNR")) |>
  group_by(method_db)  |>
  pivot_wider(names_from = metrics, values_from = values) |>
  summarise(
    TPR_mean = mean(TPR),
    TNR_mean = mean(TNR),
    TPR_sd = var(TPR),
    TNR_sd = var(TNR)
  ) |>
  mutate(bootstrap = gsub(".*__", "", method_db)) |>
  mutate(algo = gsub("__.*", "", method_db)) |>
  mutate(db = gsub(".*__", "", gsub("___.*", "", method_db))) |>
  mutate(db_unite = ifelse(grepl("Unite",method_db), "Unite", "Euk")) |>
  ggplot(aes(x = TPR_mean, y = TNR_mean, color=algo)) +
  geom_point() 

p +
  geom_segment(aes(x= TPR_mean, y = TNR_mean - TNR_sd, yend = TNR_mean + TNR_sd), alpha=0.5) +
  geom_segment(aes(y= TNR_mean, x = TPR_mean - TPR_sd, xend = TPR_mean + TPR_sd), alpha=0.5) +  ggrepel::geom_label_repel(data=subset(p$data, grepl("consensus", method_db) | TNR_mean > 0.85 | TPR_mean > 0.85), aes(label = method_db), size=2, max.overlaps = 10, force = 3) +
  xlim(c(0.5, 1)) + ylim(c(0.5, 1)) + facet_wrap(db_unite~.)
  
```


```{r}
tc_bar(d_all_taxo,
       rank_1 = "Family_sintax__mini_EUK_ITS_v1_9_3___0.5",
       rank_2 = "Family_sintax__mini_Unite___0.5",
       color_rank = "Genus_sintax__mini_EUK_ITS_v1_9_3___0.5")
```

```{r}
tc_circle(
   d_all_taxo,
  "Family_sintax__mini_Unite___0.5",
 "Family_sintax__mini_EUK_ITS_v1_9_3___0.5")
```


# Publication figures

```{r}
dir.create(here("figures"), showWarnings = FALSE)
tax_order <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Gen_sp")
```

## Q1.2 — Per-rank × method (ACC, F1, MCC)

Comparing the four methods at bootstrap 0.5; blastn shown with `rel_majority` vote
(the default; see Q1.4 for the vote-algorithm sweep).

```{r}
fig_q1_data <- res_comp_tax |>
  dplyr::filter(metrics %in% c("ACC", "F1_score", "MCC")) |>
  dplyr::filter(algo %in% c("dada2", "sintax", "lca", "blastn")) |>
  dplyr::filter(!startsWith(db, "mini_")) |>
  dplyr::filter(
    (algo %in% c("dada2", "sintax", "lca") & bootstrap_num == 0.5) |
    (algo == "blastn" & vote_algorithm == "rel_majority")
  ) |>
  dplyr::mutate(tax_level = factor(tax_level, levels = tax_order))

fig_q1_methods <- fig_q1_data |>
  ggplot(aes(x = algo, y = values, fill = db)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_grid(metrics ~ tax_level) +
  labs(x = "Method", y = "Value", fill = "Database") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(here("figures/fig_q1_methods.pdf"), fig_q1_methods, width = 16, height = 7)
ggsave(here("figures/fig_q1_methods.png"), fig_q1_methods, width = 16, height = 7, dpi = 300)
fig_q1_methods
```

## Q1.3 — Bootstrap sweep (dada2 & sintax)

F1 and MCC across `min_bootstrap ∈ {0.4, 0.5, 0.6}` per rank.

```{r}
fig_q1_bootstrap_data <- res_comp_tax |>
  dplyr::filter(metrics %in% c("F1_score", "MCC")) |>
  dplyr::filter(algo %in% c("dada2", "sintax")) |>
  dplyr::filter(!startsWith(db, "mini_")) |>
  dplyr::mutate(tax_level = factor(tax_level, levels = tax_order))

fig_q1_bootstrap <- fig_q1_bootstrap_data |>
  ggplot(aes(x = factor(bootstrap_num), y = values, fill = db)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_grid(metrics + algo ~ tax_level) +
  labs(x = "Min bootstrap", y = "Value", fill = "Database") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(here("figures/fig_q1_bootstrap.pdf"), fig_q1_bootstrap, width = 16, height = 9)
ggsave(here("figures/fig_q1_bootstrap.png"), fig_q1_bootstrap, width = 16, height = 9, dpi = 300)
fig_q1_bootstrap
```

## Q1.4 — Blastn vote-algorithm sweep

F1 and MCC across `c("rel_majority", "abs_majority", "unanimity")` per rank × DB.

```{r}
fig_q1_blastn_data <- res_comp_tax |>
  dplyr::filter(metrics %in% c("F1_score", "MCC")) |>
  dplyr::filter(algo == "blastn") |>
  dplyr::filter(!startsWith(db, "mini_")) |>
  dplyr::mutate(tax_level = factor(tax_level, levels = tax_order))

fig_q1_blastn <- fig_q1_blastn_data |>
  ggplot(aes(x = vote_algorithm, y = values, fill = db)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_grid(metrics ~ tax_level) +
  labs(x = "Vote algorithm", y = "Value", fill = "Database") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(here("figures/fig_q1_blastn_vote.pdf"), fig_q1_blastn, width = 16, height = 7)
ggsave(here("figures/fig_q1_blastn_vote.png"), fig_q1_blastn, width = 16, height = 7, dpi = 300)
fig_q1_blastn
```

## Q3.2 — Consensus vs single-method (F1, MCC)

Conservatism gradient: preference < rel_majority < abs_majority < consensus < unanimity.
Single-method bars show mean ± SE across databases.

```{r}
conservatism_order <- c(
  "dada2", "sintax", "lca", "blastn",
  "preference", "rel_majority", "abs_majority", "consensus", "unanimity"
)

fig_q3_data <- dplyr::bind_rows(
  res_comp_tax |>
    dplyr::filter(metrics %in% c("F1_score", "MCC")) |>
    dplyr::filter(algo %in% c("dada2", "sintax", "lca", "blastn")) |>
    dplyr::filter(!startsWith(db, "mini_")) |>
    dplyr::filter(
      (algo %in% c("dada2", "sintax", "lca") & bootstrap_num == 0.5) |
      (algo == "blastn" & vote_algorithm == "rel_majority")
    ) |>
    dplyr::mutate(display_method = algo, type = "single"),
  res_comp_tax |>
    dplyr::filter(metrics %in% c("F1_score", "MCC")) |>
    dplyr::filter(algo %in% paste0(c("preference", "rel_majority", "abs_majority", "consensus", "unanimity"), "_consensus")) |>
    dplyr::filter(!is_consensus | !consensus_strict) |>
    dplyr::filter(nb_agree == 1L) |>
    dplyr::mutate(display_method = gsub("_consensus$", "", algo), type = "consensus")
) |>
  dplyr::mutate(
    tax_level      = factor(tax_level, levels = tax_order),
    display_method = factor(display_method, levels = conservatism_order)
  )

fig_q3_consensus_vs_single <- fig_q3_data |>
  ggplot(aes(x = display_method, y = values, fill = type)) +
  stat_summary(fun = mean, geom = "bar", position = position_dodge()) +
  stat_summary(fun.data = mean_se, geom = "errorbar",
               position = position_dodge(width = 0.9), width = 0.25) +
  facet_grid(metrics ~ tax_level) +
  labs(x = "Method", y = "Value (mean ± SE across DBs)", fill = "Type") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(here("figures/fig_q3_consensus_vs_single.pdf"), fig_q3_consensus_vs_single, width = 16, height = 7)
ggsave(here("figures/fig_q3_consensus_vs_single.png"), fig_q3_consensus_vs_single, width = 16, height = 7, dpi = 300)
fig_q3_consensus_vs_single
```

## Q3.3 — rel_majority nb_agree_threshold sweep (1, 2, 3)

Trade-off: higher threshold → fewer FP but more FN. Faceted by rank.

```{r}
fig_q3_nb_data <- res_comp_tax |>
  dplyr::filter(metrics %in% c("F1_score", "MCC")) |>
  dplyr::filter(algo %in% c(
    "rel_majority_consensus",
    "rel_majority_nb2_consensus",
    "rel_majority_nb3_consensus"
  )) |>
  dplyr::mutate(tax_level = factor(tax_level, levels = tax_order))

fig_q3_nb_threshold <- fig_q3_nb_data |>
  ggplot(aes(x = factor(nb_agree), y = values)) +
  geom_col() +
  facet_grid(metrics ~ tax_level) +
  labs(x = "nb_agree_threshold", y = "Value") +
  theme_bw()

ggsave(here("figures/fig_q3_nb_threshold.pdf"), fig_q3_nb_threshold, width = 16, height = 7)
ggsave(here("figures/fig_q3_nb_threshold.png"), fig_q3_nb_threshold, width = 16, height = 7, dpi = 300)
fig_q3_nb_threshold
```

## M2.1 — Accuracy vs cost (F1 × wall time)

The manuscript's practical recommendation: which method gives the best accuracy per
unit of compute? Default settings only (bootstrap 0.5, blastn = rel_majority).

```{r}
benchmark_costs <- tar_read(benchmark_costs, store = "store_assign_taxo")

# F1 at every rank for default settings
f1_all_ranks <- res_comp_tax |>
  dplyr::filter(metrics == "F1_score") |>
  dplyr::filter(algo %in% c("dada2", "sintax", "lca", "blastn")) |>
  dplyr::filter(!startsWith(db, "mini_")) |>
  dplyr::filter(
    (algo %in% c("dada2", "sintax", "lca") & bootstrap_num == 0.5) |
    (algo == "blastn" & vote_algorithm == "rel_majority")
  ) |>
  dplyr::mutate(tax_level = factor(tax_level, levels = tax_order))

accuracy_vs_cost <- f1_all_ranks |>
  dplyr::left_join(
    benchmark_costs |> dplyr::select(phase, wall_time_s, peak_resident_mb, mean_cpu_pct),
    by = c("method_db" = "phase")
  )

# Species-rank panel (primary figure)
fig_m2_cost_species <- accuracy_vs_cost |>
  dplyr::filter(tax_level == "Gen_sp", !is.na(wall_time_s)) |>
  ggplot(aes(x = wall_time_s / 3600, y = values, color = algo)) +
  geom_point(size = 3, alpha = 0.85) +
  ggrepel::geom_label_repel(aes(label = db), size = 2.2, max.overlaps = 20,
                             show.legend = FALSE) +
  labs(x = "Wall time (hours)", y = "F1 score (Gen_sp rank)", color = "Method") +
  theme_bw()

ggsave(here("figures/fig_m2_cost_species.pdf"), fig_m2_cost_species, width = 8, height = 6)
ggsave(here("figures/fig_m2_cost_species.png"), fig_m2_cost_species, width = 8, height = 6, dpi = 300)
fig_m2_cost_species
```

```{r}
# All-rank panel (supplementary material)
fig_m2_cost_allranks <- accuracy_vs_cost |>
  dplyr::filter(!is.na(wall_time_s)) |>
  ggplot(aes(x = wall_time_s / 3600, y = values, color = algo, shape = db)) +
  geom_point(size = 2.5, alpha = 0.8) +
  facet_wrap(~tax_level, nrow = 1) +
  labs(x = "Wall time (hours)", y = "F1 score", color = "Method", shape = "Database") +
  theme_bw() +
  theme(legend.position = "bottom")

ggsave(here("figures/fig_m2_cost_allranks.pdf"), fig_m2_cost_allranks, width = 18, height = 5)
ggsave(here("figures/fig_m2_cost_allranks.png"), fig_m2_cost_allranks, width = 18, height = 5, dpi = 300)
fig_m2_cost_allranks
```

## M2.2 — CO₂eq per assignment target

Carbon footprint (gCO2eq) estimated via `greenAlgoR::ga_footprint()` from
`wall_time_s` in `benchmark_costs`. Hardware parameters default to the
machine's detected RAM and WORLD carbon intensity; `n_cores` is taken from
`n_threads` (set in `config.R`).

```{r}
benchmark_costs <- benchmark_costs |>
  dplyr::rowwise() |>
  dplyr::mutate(
    co2eq_gCO2 = greenAlgoR::ga_footprint(
      runtime_h      = wall_time_s / 3600,
      n_cores        = n_threads,
      add_ref_values = FALSE
    )$carbon_footprint_total_gCO2
  ) |>
  dplyr::ungroup()
```

```{r}
accuracy_vs_co2 <- f1_all_ranks |>
  dplyr::left_join(
    benchmark_costs |> dplyr::select(phase, co2eq_gCO2),
    by = c("method_db" = "phase")
  )

fig_m2_co2_species <- accuracy_vs_co2 |>
  dplyr::filter(tax_level == "Gen_sp", !is.na(co2eq_gCO2)) |>
  ggplot(aes(x = co2eq_gCO2, y = values, color = algo)) +
  geom_point(size = 3, alpha = 0.85) +
  ggrepel::geom_label_repel(
    aes(label = db), size = 2.2, max.overlaps = 20, show.legend = FALSE
  ) +
  labs(
    x = expression(CO[2]*"eq (gCO2)"),
    y = "F1 score (Gen_sp rank)",
    color = "Method"
  ) +
  theme_bw()

ggsave(here("figures/fig_m2_co2_species.pdf"), fig_m2_co2_species, width = 8, height = 6)
ggsave(here("figures/fig_m2_co2_species.png"), fig_m2_co2_species, width = 8, height = 6, dpi = 300)
fig_m2_co2_species
```

```{r}
fig_m2_co2_allranks <- accuracy_vs_co2 |>
  dplyr::filter(!is.na(co2eq_gCO2)) |>
  ggplot(aes(x = co2eq_gCO2, y = values, color = algo, shape = db)) +
  geom_point(size = 2.5, alpha = 0.8) +
  facet_wrap(~tax_level, nrow = 1) +
  labs(
    x = expression(CO[2]*"eq (gCO2)"),
    y = "F1 score",
    color = "Method",
    shape = "Database"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

ggsave(here("figures/fig_m2_co2_allranks.pdf"), fig_m2_co2_allranks, width = 18, height = 5)
ggsave(here("figures/fig_m2_co2_allranks.png"), fig_m2_co2_allranks, width = 18, height = 5, dpi = 300)
fig_m2_co2_allranks
```

## Q2.1 — Per-DB accuracy

Default settings (bootstrap 0.5 / rel_majority). Update `method_for_q2` after Q1.2
reveals the best-performing method.

```{r}
# Metadata: which simplification step does each DB represent?
db_meta <- tibble::tribble(
  ~db,                          ~db_base,   ~simplification,
  "Unite",                      "Unite",    "full",
  "Unite_Fungi",                "Unite",    "Fungi",
  "EUK_ITS_v1_9_3",             "EUK_ITS",  "full",
  "EUK_ITS_v1_9_3_Fungi",       "EUK_ITS",  "Fungi",
  "EUK_ITS_v1_9_3_Fungi_cut",   "EUK_ITS",  "Fungi+cut",
  "EUK_SSU_v1_9_3",             "EUK_SSU",  "full",
  "EUK_SSU_v1_9_3_Fungi",       "EUK_SSU",  "Fungi",
  "EUK_SSU_v1_9_3_cut",         "EUK_SSU",  "cut",
  "EUK_SSU_v1_9_3_Fungi_cut",   "EUK_SSU",  "Fungi+cut"
)

fig_q2_base <- res_comp_tax |>
  dplyr::filter(metrics %in% c("F1_score", "MCC")) |>
  dplyr::filter(algo %in% c("dada2", "sintax", "lca", "blastn")) |>
  dplyr::filter(!startsWith(db, "mini_")) |>
  dplyr::filter(
    (algo %in% c("dada2", "sintax", "lca") & bootstrap_num == 0.5) |
    (algo == "blastn" & vote_algorithm == "rel_majority")
  ) |>
  dplyr::left_join(db_meta, by = "db") |>
  dplyr::mutate(
    tax_level      = factor(tax_level,      levels = tax_order),
    simplification = factor(simplification, levels = c("full", "Fungi", "cut", "Fungi+cut"))
  )

# All methods — faceted by algo so the user can pick the best one for Q1.2
fig_q2_db <- fig_q2_base |>
  ggplot(aes(x = db, y = values, fill = simplification)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_grid(metrics + algo ~ tax_level) +
  labs(x = "Database", y = "Value", fill = "Simplification") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1))

ggsave(here("figures/fig_q2_db.pdf"), fig_q2_db, width = 18, height = 14)
ggsave(here("figures/fig_q2_db.png"), fig_q2_db, width = 18, height = 14, dpi = 300)
fig_q2_db
```

## Q2.2 — Simplification effect (paired Δ F1/MCC)

Three paired comparisons: (a) full→Fungi filter, (b) EUK_SSU full→Fungi,
(c) EUK_SSU_Fungi → primer-trimmed.

```{r}
# Helper: signed delta (simplified − full)
db_pair_delta <- function(data, db_full, db_simp, label) {
  data |>
    dplyr::filter(db %in% c(db_full, db_simp)) |>
    dplyr::select(algo, tax_level, metrics, db, values) |>
    tidyr::pivot_wider(names_from = db, values_from = values) |>
    dplyr::mutate(
      delta      = .data[[db_simp]] - .data[[db_full]],
      comparison = label
    ) |>
    dplyr::select(algo, tax_level, metrics, comparison, delta)
}

tab_q2_simplif <- dplyr::bind_rows(
  db_pair_delta(fig_q2_base, "Unite",                   "Unite_Fungi",               "(a) Unite → Fungi"),
  db_pair_delta(fig_q2_base, "EUK_ITS_v1_9_3",          "EUK_ITS_v1_9_3_Fungi",     "(b) EUK_ITS → Fungi"),
  db_pair_delta(fig_q2_base, "EUK_ITS_v1_9_3_Fungi",    "EUK_ITS_v1_9_3_Fungi_cut", "(c) EUK_ITS_Fungi → cut"),
  db_pair_delta(fig_q2_base, "EUK_SSU_v1_9_3",          "EUK_SSU_v1_9_3_Fungi",     "(d) EUK_SSU → Fungi"),
  db_pair_delta(fig_q2_base, "EUK_SSU_v1_9_3_Fungi",    "EUK_SSU_v1_9_3_Fungi_cut", "(e) EUK_SSU_Fungi → cut")
) |>
  dplyr::mutate(tax_level = factor(tax_level, levels = tax_order))

# Summary table: mean Δ per (comparison, algo, rank, metric)
tab_q2_simplif |>
  dplyr::group_by(comparison, algo, metrics, tax_level) |>
  dplyr::summarise(delta_mean = round(mean(delta, na.rm = TRUE), 3), .groups = "drop") |>
  tidyr::pivot_wider(names_from = tax_level, values_from = delta_mean) |>
  dplyr::arrange(comparison, metrics, algo) |>
  knitr::kable(caption = "Δ F1/MCC (simplified − full). Positive = simplification helps.")
```

```{r}
# Visual version of the simplification-effect table
fig_q2_simplif <- tab_q2_simplif |>
  ggplot(aes(x = algo, y = delta, fill = comparison)) +
  geom_hline(yintercept = 0, linewidth = 0.4, linetype = "dashed") +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_grid(metrics ~ tax_level) +
  labs(x = "Method", y = "Δ (simplified − full)", fill = "Comparison") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(here("figures/fig_q2_simplification.pdf"), fig_q2_simplif, width = 18, height = 8)
ggsave(here("figures/fig_q2_simplification.png"), fig_q2_simplif, width = 18, height = 8, dpi = 300)
fig_q2_simplif
```

## Q2.5 — DB size vs compute cost and accuracy

Does a smaller (filtered / trimmed) DB pay for itself in runtime without losing F1?

```{r}
db_size_tbl <- benchmark_costs |>
  dplyr::filter(!startsWith(db, "mini_")) |>
  dplyr::mutate(db_size_mb = file.size(db_path) / 1e6) |>
  dplyr::group_by(db, db_size_mb) |>
  dplyr::summarise(mean_wall_h = mean(wall_time_s, na.rm = TRUE) / 3600,
                   .groups = "drop")

f1_by_db <- res_comp_tax |>
  dplyr::filter(metrics == "F1_score", tax_level == "Gen_sp") |>
  dplyr::filter(algo %in% c("dada2", "sintax", "lca", "blastn")) |>
  dplyr::filter(!startsWith(db, "mini_")) |>
  dplyr::filter(
    (algo %in% c("dada2", "sintax", "lca") & bootstrap_num == 0.5) |
    (algo == "blastn" & vote_algorithm == "rel_majority")
  ) |>
  dplyr::group_by(db) |>
  dplyr::summarise(mean_f1 = mean(values, na.rm = TRUE), .groups = "drop")

q2_5_data <- dplyr::left_join(db_size_tbl, f1_by_db, by = "db") |>
  dplyr::left_join(db_meta |> dplyr::select(db, db_base, simplification), by = "db")

p_size_time <- q2_5_data |>
  ggplot(aes(x = db_size_mb, y = mean_wall_h, color = db_base, label = simplification)) +
  geom_point(size = 3) +
  ggrepel::geom_label_repel(size = 2.5, show.legend = FALSE) +
  labs(x = "DB size (MB)", y = "Mean wall time (h)", color = "DB family") +
  theme_bw()

p_size_f1 <- q2_5_data |>
  ggplot(aes(x = db_size_mb, y = mean_f1, color = db_base, label = simplification)) +
  geom_point(size = 3) +
  ggrepel::geom_label_repel(size = 2.5, show.legend = FALSE) +
  labs(x = "DB size (MB)", y = "Mean F1 (Gen_sp)", color = "DB family") +
  theme_bw()

fig_q2_5 <- p_size_time / p_size_f1 + patchwork::plot_layout(guides = "collect")
ggsave(here("figures/fig_q2_5_size_cost.pdf"), fig_q2_5, width = 8, height = 10)
ggsave(here("figures/fig_q2_5_size_cost.png"), fig_q2_5, width = 8, height = 10, dpi = 300)
fig_q2_5
```


## Q3.5 — Method × DB heatmap justifying the preference strategy

Which (method, DB) combination does the preference strategy prefer, and is it
a good choice? The preferred cell (sintax × EUK_ITS_v1_9_3 at bootstrap 0.5)
is highlighted in red; the preference-consensus F1 is annotated for reference.

```{r}
heat_data <- res_comp_tax |>
  dplyr::filter(
    metrics == "F1_score",
    tax_level == "Gen_sp",
    algo %in% c("dada2", "sintax", "lca", "blastn"),
    !startsWith(db, "mini_"),
    (algo %in% c("dada2", "sintax", "lca") & bootstrap_num == 0.5) |
      (algo == "blastn" & vote_algorithm == "rel_majority")
  ) |>
  dplyr::mutate(
    is_preferred = algo == "sintax" & db == "EUK_ITS_v1_9_3"
  )

pref_f1 <- res_comp_tax |>
  dplyr::filter(
    method_db == "preference_consensus",
    metrics == "F1_score",
    tax_level == "Gen_sp"
  ) |>
  dplyr::pull(values)

fig_q3_heatmap <- heat_data |>
  ggplot(aes(x = db, y = algo, fill = values)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_tile(
    data = dplyr::filter(heat_data, is_preferred),
    fill = NA, color = "red", linewidth = 1.5
  ) +
  scale_fill_viridis_c(
    "F1 (Gen_sp)", option = "plasma", direction = -1,
    limits = c(0, 1), na.value = "grey90"
  ) +
  annotate(
    "text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5, size = 3,
    label = paste0("preference consensus F1 = ", round(pref_f1, 3))
  ) +
  labs(x = "Reference database", y = "Method") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(here("figures/fig_q3_heatmap.pdf"), fig_q3_heatmap, width = 12, height = 4)
ggsave(here("figures/fig_q3_heatmap.png"), fig_q3_heatmap, width = 12, height = 4, dpi = 300)
fig_q3_heatmap
```


## Q1.6 — Per-rank method comparison table and win matrix

```{r}
method_summary_tbl <- res_comp_tax |>
  dplyr::filter(
    metrics %in% c("F1_score", "ACC"),
    algo %in% c("dada2", "sintax", "lca", "blastn"),
    !startsWith(db, "mini_"),
    (algo %in% c("dada2", "sintax", "lca") & bootstrap_num == 0.5) |
      (algo == "blastn" & vote_algorithm == "rel_majority")
  ) |>
  dplyr::group_by(tax_level, algo, metrics) |>
  dplyr::summarise(
    mean_val = round(mean(values, na.rm = TRUE), 3),
    sd_val   = round(sd(values, na.rm = TRUE), 3),
    .groups  = "drop"
  ) |>
  dplyr::mutate(mean_sd = paste0(mean_val, " (±", sd_val, ")")) |>
  tidyr::pivot_wider(
    id_cols     = c(tax_level, algo),
    names_from  = metrics,
    values_from = mean_sd
  ) |>
  dplyr::mutate(tax_level = factor(tax_level, levels = tax_order)) |>
  dplyr::arrange(tax_level, algo)

knitr::kable(
  method_summary_tbl,
  col.names = c("Rank", "Method", "ACC", "F1"),
  caption   = "Mean (±sd) F1 and ACC per method × rank (bootstrap = 0.5, averaged across DBs)"
)
```

```{r}
win_matrix_tbl <- res_comp_tax |>
  dplyr::filter(
    metrics == "F1_score",
    algo %in% c("dada2", "sintax", "lca", "blastn"),
    !startsWith(db, "mini_"),
    (algo %in% c("dada2", "sintax", "lca") & bootstrap_num == 0.5) |
      (algo == "blastn" & vote_algorithm == "rel_majority")
  ) |>
  dplyr::group_by(tax_level, db) |>
  dplyr::slice_max(values, n = 1, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::count(tax_level, algo, name = "n_wins") |>
  tidyr::pivot_wider(
    names_from  = algo,
    values_from = n_wins,
    values_fill = 0L
  ) |>
  dplyr::mutate(tax_level = factor(tax_level, levels = tax_order)) |>
  dplyr::arrange(tax_level)

knitr::kable(
  win_matrix_tbl,
  caption = "Win matrix: number of DBs where each method achieves highest F1 per rank (bootstrap = 0.5)"
)
```


## Q1.7 — Statistical test: F1 ~ method + (1|db) per rank

Linear mixed model with DB as random intercept, testing whether the method
fixed effect explains significant variance in F1 (LRT vs. null model). Requires
`lme4`.

```{r}
lmm_res_list <- lapply(
  intersect(tax_order, unique(res_comp_tax$tax_level)),
  function(rank) {
    dat <- res_comp_tax |>
      dplyr::filter(
        metrics == "F1_score",
        tax_level == rank,
        algo %in% c("dada2", "sintax", "lca", "blastn"),
        !startsWith(db, "mini_"),
        (algo %in% c("dada2", "sintax", "lca") & bootstrap_num == 0.5) |
          (algo == "blastn" & vote_algorithm == "rel_majority")
      )
    if (nrow(dat) < 8 || dplyr::n_distinct(dat$db) < 3) {
      return(NULL)
    }
    tryCatch({
      m_full <- lme4::lmer(values ~ algo + (1 | db), data = dat, REML = FALSE)
      m_null <- lme4::lmer(values ~ 1 + (1 | db), data = dat, REML = FALSE)
      lrt    <- anova(m_null, m_full)
      tibble::tibble(
        tax_level = rank,
        LRT_chisq = round(lrt$Chisq[2], 3),
        df        = as.integer(lrt[["Chi Df"]][2]),
        p_value   = round(lrt[["Pr(>Chisq)"]][2], 4),
        signif    = dplyr::case_when(
          lrt[["Pr(>Chisq)"]][2] < 0.001 ~ "***",
          lrt[["Pr(>Chisq)"]][2] < 0.01  ~ "**",
          lrt[["Pr(>Chisq)"]][2] < 0.05  ~ "*",
          TRUE ~ ""
        )
      )
    }, error = function(e) {
      tibble::tibble(tax_level = rank, LRT_chisq = NA_real_,
                     df = NA_integer_, p_value = NA_real_, signif = "ERROR")
    })
  }
)

lmm_results <- dplyr::bind_rows(lmm_res_list) |>
  dplyr::mutate(tax_level = factor(tax_level, levels = tax_order)) |>
  dplyr::arrange(tax_level)

knitr::kable(
  lmm_results,
  caption = "LRT: F1 ~ method + (1|db) vs. null (1|db) per rank. *** p<0.001, ** p<0.01, * p<0.05."
)
```


# Compute distance to true community


## En chantier

```{r}
# TODO

distinct_species_taxo_mock <- distinct(taxo_mock, Gen_sp)

mat_com_interm <- d_all_taxo |>
  psmelt() |>
  select(OTU, Sample, Abundance, contains("Gen_sp")) |>
  dplyr::filter(Abundance>0) |>
  full_join(tibble("Sample"=rep("TRUE_sample", nrow(distinct_species_taxo_mock)),
                  "Abundance"=rep(1, nrow(distinct_species_taxo_mock)),
                   "Gen_sp_true" = distinct_species_taxo_mock$Species))


mat_com <- mat_com_interm |>
  pivot_longer(contains("Gen_sp")) |>
  filter(!is.na(value)) |>
  pivot_wider(names_from = name, values_from = Abundance) |>
  mutate(across(contains("Gen_sp"), ~tidyr::replace_na(., 0))) |>
  group_by(Sample, value) |>
  summarise(across(contains("Gen_sp"), ~sum(.))) |>
  ungroup()


# mat_com |>
#   mutate(true_value = ifelse(value %in% taxo_mock$Species, 1, 0))
# 
# mat_com <- left_join(mat_com, taxo_mock$Species)
# 
# all_sp_names <- unique(c(mat_com$value, taxo_mock$Species))
# 
# 
# mat_com_list <- list()

mat_com_true <- mat_com |>
    filter(Sample=="TRUE_sample") |>
  select(value, Gen_sp_true)

mat_comA <- mat_com |> 
  filter(Sample=="samp_ITSEnz_DikA") |>
  group_by(value) |>
  select(contains("Gen_sp")) |>
#  select(contains("Unite_Fungi")) |>
#  select(contains("___0.5")) |>
  mutate(val=across(contains("Gen_sp"), ~ ifelse(.==0, 0, 1)))
  
mat_comB <- mat_com |> 
  filter(Sample=="samp_ITSEnz_DikB") |>
  group_by(value) |>
  select(contains("Gen_sp")) |>
#  select(contains("Unite_Fungi")) |>
#  select(contains("___0.5")) |>
  mutate(val=across(contains("Gen_sp"), ~ ifelse(.==0, 0, 1)))

mat_comC <- mat_com |> 
  filter(Sample=="samp_ITSEnz_DikC") |>
  group_by(value) |>
  select(contains("Gen_sp")) |>
#  select(contains("Unite_Fungi")) |>
#  select(contains("___0.5")) |>
  mutate(val=across(contains("Gen_sp"), ~ ifelse(.==0, 0, 1)))
```


## Distance between community

```{r}
compar <- full_join(mat_comA[,1:2], mat_com_true) |>
  mutate(across(everything(), ~tidyr::replace_na(., 0))) |>
  column_to_rownames("value")

vegan::vegdist(t(compar), binary=TRUE)
vegan::vegdist(t(compar), binary=FALSE)
```


```{r}
data_upset_df_comA <- full_join(mat_comA, mat_com_true, by=join_by(value==value)) |>
  column_to_rownames("value") |>
  select(contains("Gen_sp")) |>
  mutate(across(everything(), ~ ifelse(.==0, FALSE, TRUE))) |>
  as_tibble() |>
  mutate(across(everything(), ~ tidyr::replace_na(.x, FALSE))) 

data_upset_df_comB <- full_join(mat_comB, mat_com_true, by=join_by(value==value)) |>
  column_to_rownames("value") |>
  select(contains("Gen_sp")) |>
  mutate(across(everything(), ~ ifelse(.==0, FALSE, TRUE))) |>
  as_tibble() |>
  mutate(across(everything(), ~ tidyr::replace_na(.x, FALSE))) 

data_upset_df_comC <- full_join(mat_comC, mat_com_true, by=join_by(value==value)) |>
  column_to_rownames("value") |>
  select(contains("Gen_sp")) |>
  mutate(across(everything(), ~ ifelse(.==0, FALSE, TRUE))) |>
  as_tibble() |>
  mutate(across(everything(), ~ tidyr::replace_na(.x, FALSE))) 
```



```{r}
ComplexUpset::upset(data_upset_df_comA, intersect = colnames(data_upset_df_comA), min_size = 2)
ComplexUpset::upset(data_upset_df_comB, intersect = colnames(data_upset_df_comB), min_size = 2)
ComplexUpset::upset(data_upset_df_comC, intersect = colnames(data_upset_df_comC), min_size = 2)
```



###  root mean square error (RMSE)









# Compute time and cpu usage performance

```{r}
log_data_raw <- autometric::log_read("data/data_final/autometric_log_assign_taxo.txt",
                                 units_time = "hours",
                                 units_memory = "gigabytes")

conflicted::conflicts_prefer(dplyr::filter)

log_data <- log_data_raw |>
  as_tibble() |>
  filter(!grepl("conclude:", phase)) |>
  filter(!grepl("prepare:", phase)) |>
  filter(!grepl("__DEFAULT__", phase)) |>
  mutate(bootstrap = gsub(".*__", "", phase)) |>
  mutate(algo = gsub("__.*", "", phase)) |>
  mutate(db = gsub(".*__", "", gsub("___.*", "", phase)))  |>
filter(!grepl("d_asv", algo)) |>
  group_by(algo, db, bootstrap) |>
  summarise(
    phase= first(phase),
    time = n(),
    cpu = mean(cpu),
    core = mean(core),
    resident = mean(resident),
    virtual = mean(virtual)
  )
```


```{r}
log_data |> group_by(db) |> summarize(time=mean(time), cpu=mean(cpu)) |> arrange(desc(time))
log_data |> group_by(algo) |> summarize(time=mean(time), cpu=mean(cpu)) |> arrange(desc(time))
log_data |> group_by(bootstrap) |> summarize(time=mean(time), cpu=mean(cpu)) |> arrange(desc(time))
```


```{r}
log_data |>
  filter(bootstrap==0.5) |>
  ggplot(aes(x = time/3600, y = reorder(phase, dplyr::desc(time)), fill = cpu)) +
  geom_bar(stat = "identity") +
  scale_color_viridis_b("Memory (Gb)", end = 0.9, direction = -1) +
  xlab("Time (in hours)") +
  ylab("") + 
  facet_grid(db~., scales="free")


log_data |>
  ggplot(aes(x = time, y = db)) +
  geom_bar(stat="identity") +
  facet_wrap(~algo)

p_time <- log_data |>
  as_tibble() |>
  filter(!grepl("conclude:", phase)) |>
  filter(!grepl("prepare:", phase)) |>
  filter(!grepl("__DEFAULT__", phase)) |>
  filter(!grepl("d_asv", phase)) |>
  mutate(bootstrap = gsub(".*___", "", phase)) |>
  mutate(algo = gsub("__.*", "", phase)) |>
  mutate(db = gsub(".*__", "", gsub("___.*", "", phase)))  |>
  mutate(db_is_fungi = grepl("_Fungi", phase)) |>
  mutate(db_is_cut = grepl("_cut", phase)) |>
  mutate(db_base = stringr::str_remove(db, "_.*" )) |>
  group_by(phase, db, algo, db_is_fungi, db_is_cut, db_base, bootstrap) |>
  summarise(time=sum(time, na.rm = T)) |>
  ggplot(aes(x = time, y = db, fill=paste0(db_is_fungi, db_is_cut))) +
  geom_bar(stat="identity") +
  facet_grid(~algo) +
  #  geom_vline(xintercept = 16) +
  scale_fill_manual(values=c("grey", "#2585A2", "#FFA600", "#929551"))

library("patchwork")
se <- function(x) sqrt(var(x, na.rm=TRUE)/length(x))
p_time +
  (
    p_time$data |>
      group_by(algo) |>
      summarise(mean_value = mean(time, na.rm = T),
                sd_value = se(time) ) |>
      ggplot(aes(y = mean_value, x = 1)) +
      coord_flip() +
      geom_bar(stat = "identity") +
      geom_segment(aes(y=mean_value+sd_value, yend=mean_value-sd_value)) +
      facet_grid( ~ algo) +
      theme_void()
  ) +
  plot_layout(heights = c(9, 1))

p_time$data |>
  group_by(db_is_fungi, db_is_cut, db_base) |>
  summarise(mean_value = mean(time, na.rm = T),
            sd_value = se(time) ) |>
  ggplot(aes(y = mean_value, x = db_base, fill = db_base)) +
  geom_bar(stat = "identity") +
  #geom_segment(aes(y=mean_value+sd_value, yend=mean_value-sd_value)) +
  facet_grid(db_is_fungi ~ db_is_cut)

p_time$data |>
  filter(!db_is_cut) |>
  filter(!db_is_fungi) |>
  filter(algo != "idtaxa") |>
  group_by(db_base, bootstrap, algo) |>
  summarise(mean_value = mean(time, na.rm = T),
            sd_value = se(time) ) |>
  ggplot(aes(y = mean_value, x = bootstrap, fill = bootstrap)) +
  geom_bar(stat = "identity") +
  #geom_segment(aes(y=mean_value+sd_value, yend=mean_value-sd_value)) +
  facet_grid(db_base ~ algo)
```





```{r}
p_cpu <- log_data |>
  as_tibble() |>
  filter(!grepl("conclude:", phase)) |>
  filter(!grepl("prepare:", phase)) |>
  filter(!grepl("__DEFAULT__", phase)) |>
  filter(!grepl("d_asv", phase)) |>
  mutate(bootstrap = gsub(".*__", "", phase)) |>
  mutate(algo = gsub("__.*", "", phase)) |>
  mutate(db = gsub(".*__", "", gsub("___.*", "", phase)))  |>
  mutate(db_is_fungi = grepl("_Fungi", phase)) |>
  mutate(db_is_cut = grepl("_cut", phase)) |>
  mutate(db_base = stringr::str_remove(stringr::str_remove(phase, ".*__"), "_.*" )) |>
  group_by(phase, db, algo, db_is_fungi, db_is_cut, db_base) |>
  summarise(cpu=mean(cpu, na.rm = T)) |>
  ggplot(aes(x = cpu, y = db, fill=paste0(db_is_fungi, db_is_cut))) +
  geom_bar(stat="identity") +
  facet_grid(~algo, scales="free_x") +
#  geom_vline(xintercept = 16) +
  scale_fill_manual(values=c("grey", "#2585A2", "#FFA600", "#929551"))

library("patchwork")
se <- function(x) sqrt(var(x, na.rm=TRUE)/length(x))
p_cpu +
  (
    p_cpu$data |>
      group_by(algo) |>
      summarise(mean_value = mean(cpu, na.rm = T),
                sd_value = se(cpu) ) |>
      ggplot(aes(y = mean_value, x = 1)) +
      coord_flip() +
      geom_bar(stat = "identity") +
      geom_segment(aes(y=mean_value+sd_value, yend=mean_value-sd_value)) +
      facet_grid( ~ algo, scales="free_x") +
      theme_void()
  ) +
  plot_layout(heights = c(9, 1))


p_cpu$data |>
  group_by(db_is_fungi, db_is_cut, db_base,algo) |>
  summarise(mean_value = mean(cpu, na.rm = T),
            sd_value = se(cpu) ) |>
  ggplot(aes(y = log1p(mean_value), x = db_base, fill = algo, color=db_base)) +
  geom_bar(stat = "identity",position = "dodge", size=1) +
  #geom_segment(aes(y=mean_value+sd_value, yend=mean_value-sd_value)) +
  facet_grid(db_is_fungi ~ db_is_cut) +
  scale_color_manual(values=c("bisque4", "cyan4"))

```





https://adriancorrendo.github.io/metrica/articles/classification_case.html

```{r}
?metrica::confusion_matrix

```




