---
title: "In silico simulation (D1b, InSilicoSeq)"
author: "Adrien Taudière"
format: html
---

```{r setup}
#| message: false
library("conflicted")
library("MiscMetabar")
library("targets")
library("here")
library("tibble")
library("tidyr")
library("dplyr")
library("ggplot2")
library("Biostrings")

conflicted::conflicts_prefer(dplyr::filter)

here::i_am("analysis/in_silico_simulation.qmd")
source(here("config.R"))
source(here("R/load_pqverse.R"))
load_pqverse(c("MiscMetabar", "comparpq"))

# Smoke-test switch. config.R derives mini_db from TAR_PROJECT, which is unset
# when a notebook is rendered, so this file would always run on the full
# databases (one dada2 assignment against a full reference peaks at 23-45 GB,
# see docs/diagnostics_cv_and_databases.md). The chapters use
# BENCHMARK_ANALYSIS_MINI instead (analysis/00_setup.R); honour it here too:
#   BENCHMARK_ANALYSIS_MINI=TRUE quarto render analysis/in_silico_simulation.qmd
sim_mini_env <- Sys.getenv("BENCHMARK_ANALYSIS_MINI", unset = NA)
if (is.na(sim_mini_env)) {
  sim_mini <- mini_db
} else {
  sim_mini <- as.logical(sim_mini_env)
  if (is.na(sim_mini)) {
    # An unparseable value (1, yes, "true " with a trailing space) yields NA.
    # Falling back here would send the render to the full databases, the 45 GB
    # path, when mini was meant. analysis/00_setup.R stops the same way.
    stop("BENCHMARK_ANALYSIS_MINI must be TRUE or FALSE.")
  }
}
message("in silico: ", if (sim_mini) "MINI databases" else "full databases")
```

# Overview

D1b follows the Bokulich (2020) multi-dataset approach with one branch:
**InSilicoSeq** generates Illumina reads from real records of
`config.R::insilico_source`, which go through the ordinary DADA2 and assignment
pipelines. The ground truth is the taxonomy of the input records, so every ASV
can be scored per unit exactly as a mock community is.

A second branch (**D1b-B**, community matrices from miaSim's neutral and
niche-based models) existed until 2026-09-22 and was **dropped**: the benchmark
tests the taxonomic assignment of Fungi and nothing else, so simulated
community structure is out of scope (`docs/objectives_design.md` decision 12,
ROADMAP 0.9). Its code is in the git history of this file.

---

# D1b-A — InSilicoSeq path

This branch requires InSilicoSeq installed in the `cutadaptenv` conda environment
(or a dedicated env). It is not run inline — follow the steps below, then use the
DADA2 + assignment pipeline on the resulting fastqs.

## Step 1 — Prepare input fasta

Select a curated set of Fungi sequences (50–200 species, spanning ≥ 4 phyla)
from the general release named by `config.R::insilico_source`
(`Unite_s_all_20250219`, objectives_design decision 5), keeping the records
whose kingdom is Fungi. The release is used rather than its dada2-format
derivation because the truth table is parsed from the `k__`/`p__` prefixes,
which `derive_dada2()` strips (see docs/reference_databases.md); `sources/` has
no `mini_*` variant.

```{r prepare-iss-input}
# Output path for the curated input fasta
iss_input_fasta <- here("data/data_raw/insilico/iss_input_fungi.fasta")
dir.create(dirname(iss_input_fasta), showWarnings = FALSE, recursive = TRUE)

if (!file.exists(iss_input_fasta)) {
  # The general UNITE release, not the dada2-format derivation:
  # parse_unite_taxonomy() below reads the k__/p__ prefixes, which
  # derive_dada2() strips (checked 2026-09-16: 100 % of the source headers
  # carry k__, 0 % of the dada2-format ones). Using the dada2 file here would
  # give an all-NA truth table.
  db_path <- here("data/data_raw/refseq/sources", paste0(insilico_source, ".fasta"))
  all_seqs <- Biostrings::readDNAStringSet(db_path)
  all_seqs <- all_seqs[grepl("k__Fungi", names(all_seqs), fixed = TRUE)]
  set.seed(targets_seed)
  n_iss <- min(100, length(all_seqs))
  picks <- sample(length(all_seqs), n_iss)
  Biostrings::writeXStringSet(all_seqs[picks], iss_input_fasta)
  message("Wrote ", n_iss, " Fungi sequences to ", iss_input_fasta)
}
```

## Step 2 — Run InSilicoSeq (terminal)

Run from the project root. InSilicoSeq is installed in none of the conda
environments of this machine (checked 2026-09-16), but Docker is available, so
the biocontainer image is used and nothing has to be installed. Repeat with 3
different `--seed` values for replicate communities.

`--user` keeps the generated fastqs owned by you instead of root; `-v` mounts
the project root at `/mnt`, so every path inside the container is prefixed
with `/mnt`.

```bash
for seed in 1 2 3; do
  docker run --rm \
    --user "$(id -u):$(id -g)" \
    -v "$PWD":/mnt \
    quay.io/biocontainers/insilicoseq:2.0.0--pyh7cba7a3_0 \
    iss generate \
      --genomes /mnt/data/data_raw/insilico/iss_input_fungi.fasta \
      --sequence_type amplicon \
      --n_reads 10000 \
      --abundance zero_inflated_lognormal \
      --model MiSeq \
      --seed ${seed} \
      --output /mnt/data/data_raw/insilico/iss_seed${seed} \
      --cpus 4
done
```

If you would rather install it locally, `conda create -n issenv -c conda-forge
-c bioconda insilicoseq` then drop the `docker run` wrapper and the `/mnt`
prefixes.

Outputs: `data/data_raw/insilico/iss_seed{1,2,3}_R{1,2}.fastq.gz`

## Step 3 — Process with existing DADA2 + assignment pipeline

Once fastqs exist, add two projects to `_targets.yaml` (`dada2_insilico`,
`assign_taxo_insilico`) reusing `pipelines/dada2.R` and
`pipelines/assign_taxo.R` with dedicated stores, point the raw-read folder
and sample metadata at `data/data_raw/insilico/` (a per-dataset switch in
`config.R` is still to be designed, see ROADMAP D1d), then:

```r
Sys.setenv(TAR_PROJECT = "dada2_insilico");       targets::tar_make()
Sys.setenv(TAR_PROJECT = "assign_taxo_insilico"); targets::tar_make()
```

## Step 4 — Truth table

The ground truth of D1b-A is the taxonomy of the records InSilicoSeq generated
the reads from, parsed out of the UNITE headers of `iss_input_fasta`. It gives
one lineage per input record, which is the per-unit truth the scoring of
`comparpq::tc_metrics_unit()` takes (`R/score_assignments.R`, chapter 01);
matching the ASVs back to the input records follows the rule of
`R/mock_truth.R::sanger_matches()`, the input fasta playing the part of the
Sanger reads.


```{r parse-taxonomy-fn}
# Extracts the k__/p__/.../s__ taxonomy string from UNITE general-release headers.
# Headers: "ID|acc|SH|type|k__Kingdom;p__Phylum;...;s__Species"
parse_unite_taxonomy <- function(headers) {
  tax_str <- sub(".*\\|", "", headers)
  ranks   <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
  prfx    <- c("k__",     "p__",    "c__",   "o__",   "f__",    "g__",   "s__")

  result <- lapply(tax_str, function(h) {
    parts <- strsplit(h, ";")[[1]]
    vapply(seq_along(ranks), function(i) {
      hit <- grep(paste0("^", prfx[i]), parts, value = TRUE)
      if (length(hit) == 0) { return(NA_character_) }
      sub(prfx[i], "", hit[1])
    }, character(1))
  })

  df <- as.data.frame(do.call(rbind, result), stringsAsFactors = FALSE)
  names(df)    <- ranks
  rownames(df) <- headers
  df
}
```


```{r truth-table}
input_seqs <- Biostrings::readDNAStringSet(iss_input_fasta)
insilico_truth <- parse_unite_taxonomy(names(input_seqs))
head(insilico_truth)
```
