---
title: "In silico simulations (D1b)"
author: "Adrien Taudière"
format: html
---

```{r setup}
#| message: false
library("conflicted")
library("MiscMetabar")
library("targets")
library("here")
library("tibble")
library("tidyr")
library("dplyr")
library("ggplot2")
library("Biostrings")

conflicted::conflicts_prefer(dplyr::filter)

here::i_am("analysis/in_silico_simulation.qmd")
source(here("config.R"))
source(here("R/load_pqverse.R"))
load_pqverse(c("MiscMetabar", "comparpq"))

# Smoke-test switch. config.R derives mini_db from TAR_PROJECT, which is unset
# when a notebook is rendered, so this file would always run on the full
# databases (one dada2 assignment against a full reference peaks at 23-45 GB,
# see docs/diagnostics_cv_and_databases.md). The chapters use
# BENCHMARK_ANALYSIS_MINI instead (analysis/00_setup.R); honour it here too:
#   BENCHMARK_ANALYSIS_MINI=TRUE quarto render analysis/in_silico_simulation.qmd
sim_mini <- as.logical(Sys.getenv("BENCHMARK_ANALYSIS_MINI", unset = NA))
if (is.na(sim_mini)) {
  sim_mini <- mini_db
}
message("in silico: ", if (sim_mini) "MINI databases" else "full databases")

# miaSim is needed by the D1b-B branch only, and its two calls are namespaced
# (miaSim::simulateHubbell, SummarizedExperiment::assay). It is therefore not
# attached: a missing package used to abort the whole file at this chunk,
# including the D1b-A branch, which never touches it.
miasim_ok <- requireNamespace("miaSim", quietly = TRUE)
if (!miasim_ok) {
  message(
    "miaSim is not installed: the D1b-B branch below cannot run. Install it with\n",
    '  BiocManager::install("miaSim")'
  )
}
```

# Overview

Two sub-branches following the Bokulich (2020) multi-dataset approach:

- **D1b-A** — InSilicoSeq: real UNITE sequences + simulated Illumina errors → fastq
  → DADA2 → assignment. Ground truth = input fasta taxonomy.
- **D1b-B** — miaSim: neutral (Hubbell) community structure applied to real UNITE
  sequences, bypassing DADA2. Ground truth = input fasta taxonomy. Tests
  assignment accuracy independent of denoising errors.

Both branches reuse `tc_metrics_mock()` machinery from `comparpq`.

---

# D1b-A — InSilicoSeq path

This branch requires InSilicoSeq installed in the `cutadaptenv` conda environment
(or a dedicated env). It is not run inline — follow the steps below, then use the
DADA2 + assignment pipeline on the resulting fastqs.

## Step 1 — Prepare input fasta

Select a curated set of Fungi sequences (50–200 species, spanning ≥ 4 phyla) from
the general UNITE release named by `fake_ref_source` (`config.R`), the same file
D1b-B draws from, keeping the records whose kingdom is Fungi. The release is
used rather than its dada2-format derivation because the truth table is parsed
from the `k__`/`p__` prefixes, which `derive_dada2()` strips
(see docs/reference_databases.md); `sources/` has no `mini_*` variant.

```{r prepare-iss-input}
# Output path for the curated input fasta
iss_input_fasta <- here("data/data_raw/insilico/iss_input_fungi.fasta")
dir.create(dirname(iss_input_fasta), showWarnings = FALSE, recursive = TRUE)

if (!file.exists(iss_input_fasta)) {
  # The general UNITE release, not the dada2-format derivation: parse_unite_taxonomy()
  # below reads the k__/p__ prefixes, which derive_dada2() strips (checked
  # 2026-09-16: 100 % of the source headers carry k__, 0 % of the dada2-format
  # ones). Using the dada2 file here would give an all-NA truth table. D1b-B
  # draws from the same file, so both branches share one parser.
  db_path  <- here("data/data_raw/refseq/sources", paste0(fake_ref_source, ".fasta"))
  all_seqs <- Biostrings::readDNAStringSet(db_path)
  all_seqs <- all_seqs[grepl("k__Fungi", names(all_seqs), fixed = TRUE)]
  set.seed(targets_seed)
  n_iss <- min(100, length(all_seqs))
  picks <- sample(length(all_seqs), n_iss)
  Biostrings::writeXStringSet(all_seqs[picks], iss_input_fasta)
  message("Wrote ", n_iss, " Fungi sequences to ", iss_input_fasta)
}
```

## Step 2 — Run InSilicoSeq (terminal)

Run from the project root. InSilicoSeq is installed in none of the conda
environments of this machine (checked 2026-09-16), but Docker is available, so
the biocontainer image is used and nothing has to be installed. Repeat with 3
different `--seed` values for replicate communities.

`--user` keeps the generated fastqs owned by you instead of root; `-v` mounts
the project root at `/mnt`, so every path inside the container is prefixed
with `/mnt`.

```bash
for seed in 1 2 3; do
  docker run --rm \
    --user "$(id -u):$(id -g)" \
    -v "$PWD":/mnt \
    quay.io/biocontainers/insilicoseq:2.0.0--pyh7cba7a3_0 \
    iss generate \
      --genomes /mnt/data/data_raw/insilico/iss_input_fungi.fasta \
      --sequence_type amplicon \
      --n_reads 10000 \
      --abundance zero_inflated_lognormal \
      --model MiSeq \
      --seed ${seed} \
      --output /mnt/data/data_raw/insilico/iss_seed${seed} \
      --cpus 4
done
```

If you would rather install it locally, `conda create -n issenv -c conda-forge
-c bioconda insilicoseq` then drop the `docker run` wrapper and the `/mnt`
prefixes.

Outputs: `data/data_raw/insilico/iss_seed{1,2,3}_R{1,2}.fastq.gz`

## Step 3 — Process with existing DADA2 + assignment pipeline

Once fastqs exist, add two projects to `_targets.yaml` (`dada2_insilico`,
`assign_taxo_insilico`) reusing `pipelines/dada2.R` and
`pipelines/assign_taxo.R` with dedicated stores, point the raw-read folder
and sample metadata at `data/data_raw/insilico/` (a per-dataset switch in
`config.R` is still to be designed, see ROADMAP D1d), then:

```r
Sys.setenv(TAR_PROJECT = "dada2_insilico");       targets::tar_make()
Sys.setenv(TAR_PROJECT = "assign_taxo_insilico"); targets::tar_make()
```

The truth table for D1b-A is built in the same way as D1b-B below (parse taxonomy
from `iss_input_fasta` headers).

---

# D1b-B — miaSim path

Neutral (Hubbell) community dynamics applied to real UNITE sequences. No DADA2:
each simulated time-step is treated as a perfectly denoised sample.

## Helper — parse UNITE taxonomy from header

```{r parse-taxonomy-fn}
# Extracts the k__/p__/.../s__ taxonomy string from UNITE general-release headers.
# Headers: "ID|acc|SH|type|k__Kingdom;p__Phylum;...;s__Species"
parse_unite_taxonomy <- function(headers) {
  tax_str <- sub(".*\\|", "", headers)
  ranks   <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
  prfx    <- c("k__",     "p__",    "c__",   "o__",   "f__",    "g__",   "s__")

  result <- lapply(tax_str, function(h) {
    parts <- strsplit(h, ";")[[1]]
    vapply(seq_along(ranks), function(i) {
      hit <- grep(paste0("^", prfx[i]), parts, value = TRUE)
      if (length(hit) == 0) { return(NA_character_) }
      sub(prfx[i], "", hit[1])
    }, character(1))
  })

  df <- as.data.frame(do.call(rbind, result), stringsAsFactors = FALSE)
  names(df)    <- ranks
  rownames(df) <- headers
  df
}
```

## 1. Sample species from UNITE Fungi

```{r sample-species}
# General UNITE release (k__/p__ headers, see docs/reference_databases.md),
# restricted to the records whose kingdom is Fungi.
db_path <- here("data/data_raw/refseq/sources", paste0(fake_ref_source, ".fasta"))
all_fungi <- Biostrings::readDNAStringSet(db_path)
all_fungi <- all_fungi[grepl("k__Fungi;", names(all_fungi), fixed = TRUE)]

# Keep only sequences with ≥ genus-level annotation.
has_genus <- !grepl("g__.*Incertae_sedis|g__[^;]+_gen_", names(all_fungi))
all_fungi <- all_fungi[has_genus]

n_species <- min(50, length(all_fungi))
set.seed(targets_seed)
idx       <- sample(length(all_fungi), n_species)
sim_seqs  <- all_fungi[idx]

# Truth table: one row per species, columns = taxonomy ranks.
truth_table <- parse_unite_taxonomy(names(sim_seqs))

cat("Species selected:", n_species, "\n")
cat("Phyla represented:", length(unique(truth_table$Phylum)), "\n")
print(table(truth_table$Phylum))
```

## 2. Simulate community abundances with Hubbell model

Run `n_samples` independent Hubbell simulations (different seeds = independent
communities from the same metacommunity). Each returns a species × time matrix;
we take the final time step as the community composition.

```{r simulate-hubbell}
if (!miasim_ok) {
  stop(
    'miaSim is required from here on. Install it with BiocManager::install("miaSim"), ',
    "or render only the D1b-A section above."
  )
}

n_samples <- 10
sim_seed  <- targets_seed

all_counts <- vapply(seq_len(n_samples), function(s) {
  set.seed(sim_seed + s)
  sim <- miaSim::simulateHubbell(
    n_species         = n_species,
    M                 = 200,
    carrying_capacity = 1000,
    k_events          = 50,
    migration_p       = 0.02,
    t_end             = 200
  )
  counts_mat <- SummarizedExperiment::assay(sim, "counts")
  as.integer(counts_mat[, ncol(counts_mat)])
}, integer(n_species))

rownames(all_counts) <- names(sim_seqs)
colnames(all_counts) <- paste0("sim_sample_", seq_len(n_samples))

cat("OTU table dimensions:", nrow(all_counts), "species ×", ncol(all_counts), "samples\n")
cat("Samples with 0 reads:", sum(colSums(all_counts) == 0), "\n")
```

## 3. Build phyloseq + inject fake taxa

```{r build-phyloseq}
pq_sim <- phyloseq::phyloseq(
  phyloseq::otu_table(all_counts, taxa_are_rows = TRUE),
  phyloseq::sample_data(data.frame(
    sample_id = colnames(all_counts),
    row.names = colnames(all_counts)
  ))
)
pq_sim@refseq <- sim_seqs

# Inject shuffled (fake_*) and external (non-Fungi) sequences for TN computation,
# mirroring the real pipeline in script_assign_taxo_parallel.R.
pq_sim <- add_shuffle_seq_pq(pq_sim, prop_fake = prop_fake)
pq_sim <- add_external_seq_pq(
  pq_sim,
  Biostrings::readDNAStringSet(here(fake_ref_fasta))
)

cat("Phyloseq taxa after fake injection:", ntaxa(pq_sim), "\n")
```

## 4. Assign taxonomy

Run a small subset of method × DB combinations. Adjust `sim_values_map` to test
more combinations. `sim_mini` (set at the top of this file from
`BENCHMARK_ANALYSIS_MINI`, falling back to `config.R::mini_db`) controls which
database files are used: `mini_*` for smoke-testing, full for publication. Keep
it TRUE while testing — one dada2 assignment against a full reference peaks at
23–45 GB (docs/diagnostics_cv_and_databases.md).

```{r assign-taxonomy}
# Database names come from config.R (db_list, built from benchmark_dbs): never
# glue a suffix by hand (CLAUDE.md). Picking the Fungi variant out of db_list
# fails loudly if a release is renamed, instead of silently dropping the row at
# the file.exists() filter below.
preference_db_fungi <- grep(
  paste0("^", preference_db, "_Fungi$"),
  db_list,
  value = TRUE
)
stopifnot(length(preference_db_fungi) == 1)

sim_values_map_all <- tibble::tibble(
  method = c("sintax", "sintax", "dada2"),
  db     = c(seed_taxonomy_db, preference_db_fungi, seed_taxonomy_db),
  min_bootstrap = 0.5
) |>
  dplyr::mutate(
    # `if`, not ifelse(): a scalar ifelse() recycles its first value (S6.1).
    db_name   = if (sim_mini) paste0("mini_", db) else db,
    db_path   = paste0(
      ifelse(method == "dada2",
             "data/data_raw/refseq/dada2_format/",
             "data/data_raw/refseq/sintax_format/"),
      db_name, ".fasta"
    ),
    full_name = paste0(method, "__", db, "___", min_bootstrap)
  )

missing_db <- dplyr::filter(sim_values_map_all, !file.exists(here(db_path)))
if (nrow(missing_db) > 0) {
  message(
    "Database files not found, rows dropped: ",
    paste(missing_db$db_path, collapse = ", ")
  )
}

sim_values_map <- dplyr::filter(sim_values_map_all, file.exists(here(db_path)))

if (nrow(sim_values_map) == 0) {
  stop("No database files found. Run derive_all_variants() first.")
}

pq_assigned <- pq_sim
for (i in seq_len(nrow(sim_values_map))) {
  row <- sim_values_map[i, ]
  message("Assigning: ", row$full_name)
  # clean_pq = FALSE: assign_sintax() and assign_vsearch_lca() drop empty taxa
  # before querying, which would skip the fake_ / external_ controls and leave
  # no true negatives (ROADMAP B19, same fix as pipelines/assign_taxo.R).
  # add_new_taxonomy_pq() filters `...` on the formals of the called function,
  # so clean_pq reaches sintax and lca and is ignored for dada2.
  pq_assigned <- add_new_taxonomy_pq(
    pq_assigned,
    method        = row$method,
    ref_fasta     = here(row$db_path),
    suffix        = paste0("_", row$full_name),
    min_bootstrap = row$min_bootstrap,
    clean_pq      = FALSE
  )
}
```

## 5. Compute performance metrics

`tc_metrics_mock()` takes a `ranks_df` (one column per `method_db`, one row per
rank) and a `true_values_df` (one column per rank, same order), and returns the
long format `method_db` / `tax_level` / `metrics` / `values`. Every method
names its columns `<rank>_<full_name>`: dada2 through
`paste0(colnames(tax_tab), suffix)`, sintax, lca and blastn through
`paste0(taxa_ranks, suffix)`.

```{r compute-metrics}
sim_ranks <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")

ranks_df <- do.call(
  data.frame,
  stats::setNames(
    lapply(sim_values_map$full_name, \(fn) paste0(sim_ranks, "_", fn)),
    sim_values_map$full_name
  )
)

missing_cols <- setdiff(
  as.vector(unlist(ranks_df)),
  colnames(pq_assigned@tax_table)
)
if (length(missing_cols) > 0) {
  print(missing_cols)
  stop("Assignment columns expected from sim_values_map are missing.")
}

res_sim <- tc_metrics_mock(
  pq_assigned,
  ranks_df       = ranks_df,
  true_values_df = truth_table[, sim_ranks]
) |>
  dplyr::left_join(
    sim_values_map |> dplyr::select(full_name, method, db, min_bootstrap),
    by = c("method_db" = "full_name")
  )

head(res_sim)
```

## 6. Visualise results

### F1 and MCC per taxonomic rank

```{r plot-metrics}
res_sim |>
  dplyr::filter(metrics %in% c("F1_score", "MCC")) |>
  ggplot2::ggplot(ggplot2::aes(
    x    = tax_level,
    y    = values,
    fill = db
  )) +
  ggplot2::geom_col(position = "dodge") +
  ggplot2::facet_grid(metric ~ method, scales = "free_y") +
  ggplot2::labs(
    title = "D1b-B: assignment accuracy on miaSim communities",
    x     = "Taxonomic rank",
    y     = "Score",
    fill  = "Reference DB"
  ) +
  ggplot2::theme_bw()
```

### NA (unclassified) proportion per rank

`tc_metrics_mock()` returns no NA metric, so the proportion of unassigned taxa
is computed from the assignment columns themselves, over the real taxa only
(the `fake_` and `external_` controls are unassignable by construction).

```{r plot-na}
tax_df    <- as.data.frame(as(phyloseq::tax_table(pq_assigned), "matrix"))
real_taxa <- !grepl("^fake_|^external_", rownames(tax_df))

na_prop <- do.call(rbind, lapply(sim_values_map$full_name, function(fn) {
  cols <- paste0(sim_ranks, "_", fn)
  data.frame(
    method_db = fn,
    tax_level = factor(sim_ranks, levels = sim_ranks),
    prop_NA   = vapply(cols, \(cl) mean(is.na(tax_df[real_taxa, cl])), numeric(1)),
    row.names = NULL
  )
})) |>
  dplyr::left_join(
    sim_values_map |> dplyr::select(full_name, method, db),
    by = c("method_db" = "full_name")
  )

na_prop |>
  ggplot2::ggplot(ggplot2::aes(
    x      = tax_level,
    y      = prop_NA,
    colour = method,
    group  = method_db
  )) +
  ggplot2::geom_line() +
  ggplot2::geom_point() +
  ggplot2::facet_wrap(~db) +
  ggplot2::labs(
    title = "D1b-B: proportion unclassified per rank",
    x     = "Taxonomic rank",
    y     = "Proportion NA"
  ) +
  ggplot2::theme_bw()
```

---

# Comparison: D1b-B vs mock community (D1)

The mock-community metrics no longer have to be recomputed here:
`analysis/01_load_and_clean.qmd` writes `res_comp_tax` to
`analysis/_cache/res_comp_tax.rds` (render chapter 01 first). Only the six
ranks common to both tables are compared: chapter 01 replaces Species by the
concatenated `Gen_sp`, which this notebook does not build.

```{r compare-mock}
cache_file <- here("analysis/_cache/res_comp_tax.rds")

if (!file.exists(cache_file)) {
  message("Render analysis/01_load_and_clean.qmd first: ", cache_file, " is missing.")
} else {
  shared_ranks <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus")

  res_mock <- readRDS(cache_file) |>
    dplyr::filter(!is_consensus, algo %in% c("dada2", "sintax", "lca", "blastn")) |>
    dplyr::select(method_db, tax_level, metrics, values, method = algo, db)

  res_combined <- dplyr::bind_rows(
    dplyr::mutate(
      dplyr::select(res_sim, method_db, tax_level, metrics, values, method, db),
      dataset = "in_silico_miaSim"
    ),
    dplyr::mutate(res_mock, dataset = "mock_community")
  ) |>
    dplyr::filter(tax_level %in% shared_ranks) |>
    dplyr::mutate(tax_level = factor(tax_level, levels = shared_ranks))

  res_combined |>
    dplyr::filter(metrics == "F1_score") |>
    ggplot2::ggplot(ggplot2::aes(x = tax_level, y = values, fill = dataset)) +
    ggplot2::geom_boxplot() +
    ggplot2::facet_wrap(~method) +
    ggplot2::labs(
      title = "F1 per rank: in silico communities against the mock community",
      x     = "Taxonomic rank",
      y     = "F1",
      fill  = "Dataset"
    ) +
    ggplot2::theme_bw()
}
```
